---
title: "Christiane_Kraemer_report"

output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
date: "2025-02-10"
---

```{r}
library(knitr)

opts_chunk$set(
  fig.path = paste0("./figures/"),
  fig.align = "center",
  fig.show = "asis",
  eval = TRUE,
  fig.width = 10,
  fig.height = 7,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  size = "small",
  comment = "##",
  echo = TRUE,
  results = "markup",
  dev = c("png", "pdf")
)
options(
  replace.assign = TRUE,
  width = 80
)
```



# Loading the library

```{r}
library("readxl")
library("readr")
library("DESeq2")
library("tximeta")
library("SummarizedExperiment")
library("org.Hs.eg.db")
library("macrophage")
library("DESeq2")
library("clusterProfiler")
library("org.Hs.eg.db")
library("enrichplot")
library("mosdef")
library("GeneTonic")
library("IHW")
library("gplots")
library("genefilter")
library("rtracklayer")
library("xtable")
library("GO.db")
library("goseq")
library("GenomicFeatures")
library("ReactomePA")
library("dplyr")
library("NMF")
library("pheatmap")
library("GOplot")
library("knitr")
library("pcaExplorer")
library("DT")
library("ideal")
library("edgeR")
library("hexbin")
library("latticeExtra")
library("vsn")
library("gplots")
library("RColorBrewer")
library("topGO")
library("BIUMmisc")
library("biomaRt")
library("pathview")
library("VennDiagram")
library("GSEABase")
library("msigdbr")
library("ConsensusClusterPlus")
library("visNetwork")
library("igraph")
library("ggraph")
library("decoupleR")
library("OmnipathR")
library("quantiseqr")
library("immunedeconv")
library("tibble")
library("tidyr")
library("ComplexHeatmap")
library("plotly")

```

# loading the coldata and count matrix for LR

```{r}
colData <- read_excel("./coldata_phybion_pilot_LR.xlsx")

colData <- as.data.frame(colData)

count_matrix <- read_tsv("./Isoquant/OUT/OUT.gene_grouped_counts.tsv")

count_matrix <- as.matrix(count_matrix)

rownames(count_matrix) <- count_matrix[, 1]

count_matrix <- count_matrix[, -1]

colnames(count_matrix) <- colData$sample_id

rownames(colData) <- colnames(count_matrix)

gene_id <- rownames(count_matrix)

count_matrix <- apply(count_matrix, 2, as.numeric)
rownames(count_matrix) <- gene_id

```

# Creating the dds_LR object

```{r}
colData$Dose <- factor(colData$Dose)

colData$Dose <- relevel(colData$Dose, ref = "0Gy")

colData$Donors <- factor(colData$Donors)

dds_LR <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = colData,
  design = ~ Donors + Dose
)

dds_LR

dds_LR <- dds_gencode_to_ensembl(dds_LR)


dds_LR$Dose <- factor(dds_LR$Dose)

dds_LR$Dose <- relevel(dds_LR$Dose, ref = "0Gy")

```

# Creating dds_SR

```{r}

gse <- readRDS("gse_phybion_ShortReads.RDS")


dds_SR <- DESeqDataSet(gse, design = ~ Donors + Dose)

dds_SR <- dds_SR[, c(
  "0Gy-2hrs-F", "0Gy-2hrs-A", "0Gy-2hrs-S", "0Gy-6hrs-F", "0Gy-6hrs-A",
  "0Gy-6hrs-S", "0.5Gy-2hrs-F", "0.5Gy-2hrs-A", "0.5Gy-2hrs-S", "0.5Gy-6hrs-F",
  "0.5Gy-6hrs-A", "0.5Gy-6hrs-S", "1Gy-2hrs-F", "1Gy-2hrs-A", "1Gy-2hrs-S",
  "1Gy-6hrs-F", "1Gy-6hrs-A", "1Gy-6hrs-S", "2Gy-2hrs-F", "2Gy-2hrs-A",
  "2Gy-2hrs-S", "2Gy-6hrs-F", "2Gy-6hrs-A", "2Gy-6hrs-S", "4Gy-2hrs-F",
  "4Gy-2hrs-A", "4Gy-2hrs-S", "4Gy-6hrs-F", "4Gy-6hrs-A", "4Gy-6hrs-S"
)]

dds_SR <- dds_gencode_to_ensembl(dds_SR)

dds_SR_t2 <- dds_SR[, dds_SR$Time_point == "6hrs"]

dds_SR_4Gy <- dds_SR_t2[, dds_SR_t2$Dose %in% c("0Gy", "4Gy")]


dds_SR_4Gy$Dose<- droplevels(dds_SR_4Gy$Dose)


dds_SR_4Gy$Dose <- relevel(dds_SR_4Gy$Dose, "0Gy")


design(dds_SR_4Gy) <- ~ Donors + Dose

```


# Variance stablizing transformation

```{r}

vsd_phybion_LR <- vst(dds_LR)
vsd_phybion_SR <- vst(dds_SR_4Gy)


vsd_corrected_LR <- vsd_phybion_LR
vsd_corrected_SR <- vsd_phybion_SR


assay(vsd_corrected_LR) <- limma::removeBatchEffect(assay(vsd_phybion_LR), batch = vsd_phybion_LR$Donors)

assay(vsd_corrected_SR) <- limma::removeBatchEffect(assay(vsd_phybion_SR), batch = vsd_phybion_SR$Donors)


```

# Exploratory Data Anlaysis

```{r}
plot_totcounts(dds_LR, group = "Dose")

plot_totcounts(dds_SR_4Gy, group = "Dose")

```

## Histogram of data distribution

LR
```{r}
hist(assay(vsd_phybion_LR)[rowSums(assay(vsd_phybion_LR))>50,])

```

SR
```{r}
hist(assay(vsd_phybion_SR)[rowSums(assay(vsd_phybion_SR))>50,])

```

## PCA plot

```{r}

pca_plot_LR <- plotPCA(vsd_phybion_LR, intgroup = c("Donors", "Dose"), returnData = TRUE)

percentVar <- round(100 * attr(pca_plot_LR, "percentVar"))

PCA_LR<- ggplot(pca_plot_LR, aes(x = PC1, y = PC2, color = Donors)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17)) + 
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() + theme(
  plot.title = element_text(size = 17), 
  axis.title.x = element_text(size = 15), 
  axis.title.y = element_text(size = 15), 
  axis.text = element_text(size = 13),   
  legend.text = element_text(size = 13), 
  legend.title = element_text(size = 15) 
)

```

```{r}

pca_plot_SR <- plotPCA(vsd_phybion_SR, intgroup = c("Donors", "Dose"), returnData = TRUE)

percentVar <- round(100 * attr(pca_plot_SR, "percentVar"))

PCA_SR<- ggplot(pca_plot_SR, aes(x = PC1, y = PC2, color = Donors)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17)) + # Adjust the number of shapes if necessary
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()+
  theme(plot.background = element_rect(fill = "white", color = NA))
```

### PCA LR
```{r}

PCA_LR

```

### PCA LR

```{r}

PCA_SR

```

# Annotation

```{r}
anno_df <- get_annotation_orgdb(
  dds = dds_LR,
  orgdb_species = "org.Hs.eg.db",
  idtype = "ENSEMBL"
)
```

# Differential Expression Analysis---> LR

```{r}

dds_LR$Donors <- factor(dds_LR$Donors)

dds_LR$Dose <- factor(dds_LR$Dose)

design(dds_LR) <- ~ Donors + Dose

dds_LR <- DESeq(dds_LR)


resultsNames(dds_LR)

results_LR <- results(dds_LR, name = "Dose_4Gy_vs_0Gy", alpha = 0.05)

summary(results_LR)

results_LR_shrink <- lfcShrink(dds_LR, res = results_LR, coef = "Dose_4Gy_vs_0Gy", type = "apeglm")

results_LR_shrink_DEs <- deresult_to_df(results_LR_shrink, FDR = 0.05)


View(results_LR_shrink_DEs)

genes <- rownames(results_LR_shrink_DEs)

genes <- mapIds(
  org.Hs.eg.db,
  keys = genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

results_LR_shrink_DEs$SYMBOL<- genes

```

# Differential Expression Analysis---> SR

```{r}
dds_SR_4Gy <- DESeq(dds_SR_4Gy)

resultsNames(dds_SR_4Gy)

results_SR <- results(dds_SR_4Gy, name = "Dose_4Gy_vs_0Gy", alpha = 0.05)

results_SR_shrink<- lfcShrink(dds_SR_4Gy, res = results_SR, coef = "Dose_4Gy_vs_0Gy", type = "apeglm")

summary(results_SR_shrink)

results_SR_shrink_DE <- deresult_to_df(results_SR_shrink, FDR = 0.05)


genes_SR <- rownames(results_SR_shrink_DE)

genes_SR <- mapIds(
  org.Hs.eg.db,
  keys = genes_SR,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

results_SR_shrink_DE$SYMBOL<- genes_SR


# Deseq short Vs Deseq long

common_DEs <- intersect(rownames(results_SR), rownames(results_LR))

results_SR_shrink_common <- results_SR_shrink[common_DEs, ]

results_LR_shrink_common<- results_LR_shrink[common_DEs, ]


results_SR_common <- results_SR[common_DEs, ]

results_LR_common<- results_LR[common_DEs, ]


results_SR_common_df<- as.data.frame(results_SR_shrink_common)


results_LR_common_df<- as.data.frame(results_LR_shrink_common)


genes <- rownames(results_LR_common_df)

genes <- mapIds(
  org.Hs.eg.db,
  keys = genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

results_LR_common_df$SYMBOL<- genes

### unshrunken plot

plot(results_LR_common$log2FoldChange, results_SR_common$log2FoldChange, xlim = c(-10, 10), ylim = c(-10, 10))
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

### shrunken plot

plot(results_LR_shrink_common$log2FoldChange, results_SR_shrink_common$log2FoldChange, xlim = c(-10, 10), ylim = c(-10, 10))
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

### interactive plot

plot_ly(
    data = results_LR_common_df, 
    x = ~log2FoldChange, 
    y = ~results_SR_shrink_common$log2FoldChange, 
    text = ~SYMBOL,
    type = 'scatter',
    mode = 'markers'
) %>%
    layout(
        xaxis = list(title = "Log2 Fold Change (LR)"),
        yaxis = list(title = "Log2 Fold Change (SR)"),
        shapes = list(
            list(
                type = "line",
                x0 = -10, x1 = 10, y0 = -10, y1 = 10,
                line = list(color = "red", dash = "dot", width = 2)
            )
        )
    )

```

## Exploring more on pvalues
SR
```{r}
hist(results_SR_shrink_DE$pvalue)
```
LR
```{r}
hist(results_LR_shrink_DEs$pvalue)
```

```{r}
## Venndiagram of DEGs

gplots::venn(
  list(
    rownames(results_SR_shrink_DE), rownames(results_LR_shrink_DEs)
  )
)

common_DEGs<- intersect(rownames(results_SR_shrink_DE), rownames(results_LR_shrink_DEs))

common_DEGs<- mapIds(org.Hs.eg.db,
  keys = common_DEGs,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

print(common_DEGs)


## Venndiagram of all detected genes in SR and LR

gplots::venn(
  list(
    rownames(counts(dds_SR_4Gy)[rowSums(counts(dds_SR_4Gy))>0,] ), rownames(counts(dds_LR)[rowSums(counts(dds_LR))>0,])
  )
)

unique_detected_SR_genes<- setdiff(rownames(counts(dds_SR_4Gy)[rowSums(counts(dds_SR_4Gy))>0,]),rownames(counts(dds_LR)[rowSums(counts(dds_LR))>0,]))

unique_detected_LR_genes<- setdiff(rownames(counts(dds_LR)[rowSums(counts(dds_LR))>0,]), rownames(counts(dds_SR_4Gy)[rowSums(counts(dds_SR_4Gy))>0,]))




unique_detected_SR_genes_df<- as.data.frame(counts(dds_SR_4Gy)[unique_detected_SR_genes , ])

unique_detected_LR_genes_df<- as.data.frame(counts(dds_LR)[unique_detected_LR_genes , ])

unique_detected_LR_genes_df<- unique_detected_LR_genes_df[rowSums(unique_detected_LR_genes_df)>30, ]

unique_detected_SR_genes_df<- unique_detected_SR_genes_df[rowSums(unique_detected_SR_genes_df)>30, ]




unique_detected_SR_genes_Symbols <- mapIds(
  org.Hs.eg.db,
  keys = rownames(unique_detected_SR_genes_df),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)



unique_detected_LR_genes_Symbols <- mapIds(
  org.Hs.eg.db,
  keys = rownames(unique_detected_LR_genes_df),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

unique_detected_SR_genes_df$SYMBOL<- unique_detected_SR_genes_Symbols

unique_detected_LR_genes_df$SYMBOL<- unique_detected_LR_genes_Symbols

na.omit(unique_detected_LR_genes_df) |> View()

```

# Identifying the difference between SR and LR

```{r}

common_DEGs <- intersect(results_SR_shrink_DE$SYMBOL, results_LR_shrink_DEs$SYMBOL)

unique_LR<- setdiff(results_LR_shrink_DEs$SYMBOL,results_SR_shrink_DE$SYMBOL)

unique_SR<- setdiff(results_SR_shrink_DE$SYMBOL,results_LR_shrink_DEs$SYMBOL)


result_SR_4Gy_full<- readRDS("result_SR_4Gy_full.RDS")

result_SR_4Gy_full <- deresult_to_df(result_SR_4Gy_full, FDR = 0.05)


extracting_true<- unique_LR %in% result_SR_4Gy_full$SYMBOL

LR_only_genes<- unique_LR[!extracting_true]


# out of the 46 LR_only genes, extracting genes that are and are not captured in dds_SR

LR_genes_compare <- LR_only_genes %in% rowData(dds_SR)$SYMBOL

LR_DEGs_in_dds_SR<- LR_only_genes[LR_genes_compare]
LR_DEGs_absent_in_dds_SR<- LR_only_genes[!LR_genes_compare]

#LR DEGs in dds_SR count matrix
print(LR_DEGs_in_dds_SR)

#LR DEGs absent in dds_SR count matrix
print(LR_DEGs_absent_in_dds_SR)


```

# Volcano plot

```{r}

# Volcano plot LR

volcPlot_LR <- de_volcano(
  res_de = results_LR_shrink,
  mapping = "org.Hs.eg.db",
  labeled_genes = 50,
  logfc_cutoff = 1
)
volcPlot_LR

# Volcano plot SR

volcPlot_SR <- de_volcano(
  res_de = results_SR_shrink,
  mapping = "org.Hs.eg.db",
  labeled_genes = 50,
  logfc_cutoff = 1
)
volcPlot_SR

```

## Creating gene plotting function

```{r}
plot_gene <- function (dds_list, gene, intgroup = "condition", assay = "counts", 
                                annotation_obj = NULL, normalized = TRUE, transform = TRUE, 
                                labels_display = FALSE, labels_repel = TRUE, plot_type = "auto", 
                                return_data = FALSE, color_by = "Donors", plot_titles = NULL) 
{
  plot_type <- match.arg(plot_type, c("auto", "jitteronly", "boxplot", "violin", "sina"))
  
  # Ensure plot_titles matches the length of dds_list
  if (!is.null(plot_titles) && length(plot_titles) != length(dds_list)) {
    stop("The length of `plot_titles` must match the number of elements in `dds_list`.")
  }
  
  # Collect data from both dds objects
  df_list <- lapply(seq_along(dds_list), function(i) {
    dds <- dds_list[[i]]
    if (!(all(intgroup %in% colnames(colData(dds))))) {
      stop("`intgroup` not found in the colData slot of the dds object for dds ", i)
    }
    
    df <- get_expression_values(dds = dds, gene = gene, intgroup = intgroup, 
                                assay = assay, normalized = normalized)
    df$sample_id <- rownames(df)
    
    # Use custom titles if provided
    df$source <- if (!is.null(plot_titles)) {
      plot_titles[i]
    } else {
      paste0("dds_", i)  # Default identifier for the dds object
    }
    
    # Add Donors column if it exists
    if ("Donors" %in% colnames(colData(dds))) {
      df$Donors <- colData(dds)[, "Donors"]
    } else {
      stop("The `Donors` column is missing in the colData of the dds object for dds ", i)
    }
    
    return(df)
  })
  df <- bind_rows(df_list)

  # Add gene annotation if provided
  if (!is.null(annotation_obj)) {
    genesymbol <- annotation_obj$gene_name[match(gene, annotation_obj$gene_id)]
  } else {
    genesymbol <- gene
  }

  onlyfactors <- df[, match(intgroup, colnames(df))]
  df$plotby <- interaction(onlyfactors)
  min_by_groups <- min(table(df$plotby))

  if (return_data) {
    return(df)
  }

  # Create ggplot
  p <- ggplot(df, aes(x = .data$plotby, y = .data$exp_value, col = .data[[color_by]])) + 
    scale_x_discrete(name = "") + 
    scale_color_discrete(name = "Experimental\ngroup") + 
    theme_bw() + 
    facet_wrap(~source, scales = "fixed")  # Use custom titles for facets
  
  jit_pos <- position_jitter(width = 0.2, height = 0, seed = 42)
  
  # Plot individual points for samples/donors
  p <- p + geom_point(position = jit_pos, shape = 16, size = 3)

  # Display labels if required
  if (labels_display) {
    if (labels_repel) {
      p <- p + ggrepel::geom_text_repel(aes(label = .data$sample_id), 
                                        min.segment.length = 0, position = jit_pos)
    } else {
      p <- p + geom_text(aes(label = .data$sample_id), hjust = -0.1, vjust = 0.1, position = jit_pos)
    }
  }
  
  # Add a connected mean line across all groups
  p <- p + stat_summary(fun = mean, geom = "line", aes(group = 1), 
                        color = "grey80", linewidth = 0.8)

  y_label <- if (assay == "counts" & normalized) {
    "Normalized counts"
  } else if (assay == "counts" & !normalized) {
    "Counts"
  } else if (assay == "abundance") {
    "TPM - Transcripts Per Million"
  } else {
    assay
  }
  
  if (transform) {
    p <- p + scale_y_log10(name = paste0(y_label, " (log10 scale)"))
  } else {
    p <- p + scale_y_continuous(name = y_label)
  }

  p <- p + labs(title = paste0(genesymbol, " - ", gene))
  
  return(p)
}


```

## Checking some well-known radiosensitive genes expression direction

### CDKN1A

```{r}

plot_gene(
  dds= list(dds_SR_4Gy, dds_LR),
  gene = c("ENSG00000124762"),
  intgroup = "Dose",
  annotation_obj = anno_df,
  plot_titles = c("SR", "LR"))

```

### FDXR
```{r}

plot_gene(
  dds= list(dds_SR_4Gy, dds_LR),
  gene = c("ENSG00000161513"),
  intgroup = "Dose",
  annotation_obj = anno_df,
  plot_titles = c("SR", "LR"))

```

### AEN
```{r}

plot_gene(
  dds= list(dds_SR_4Gy, dds_LR),
  gene = c("ENSG00000181026"),
  intgroup = "Dose",
  annotation_obj = anno_df,
  plot_titles = c("SR", "LR"))

```

# Functional Enrichment Analysis

```{r}
expressedInAssay_LR <- (rowSums(assay(dds_LR)) > 0)
geneUniverseExprENS_LR <- rownames(dds_LR)[expressedInAssay_LR]
geneUniverseExpr_LR <- anno_df$gene_name[match(geneUniverseExprENS_LR, anno_df$gene_id)]


expressedInAssay_SR <- (rowSums(assay(dds_SR_4Gy)) > 0)
geneUniverseExprENS_SR <- rownames(dds_LR)[expressedInAssay_SR]
geneUniverseExpr_SR <- anno_df$gene_name[match(geneUniverseExprENS_SR, anno_df$gene_id)]

```

```{r}
## enrichment LR

res_enrich_LR <- run_cluPro(
  de_genes = results_LR_shrink_DEs$id,
  res_de = results_LR,
  de_container = dds_LR,
  keyType = "ENSEMBL",
  mapping = "org.Hs.eg.db",
  ont = "BP",
  pAdjustMethod = "BH",
  bg_genes = geneUniverseExpr_LR,
  FDR_threshold = 0.05,
)

sig_res_enrich_LR <- res_enrich_LR@result[res_enrich_LR@result$p.adjust < 0.05, ]

## enrichment SR

res_enrich_SR <- run_cluPro(
  de_genes = results_SR_shrink_DE$id,
  res_de = results_SR,
  de_container = dds_SR_4Gy,
  keyType = "ENSEMBL",
  mapping = "org.Hs.eg.db",
  ont = "BP",
  pAdjustMethod = "BH",
  bg_genes = geneUniverseExpr_SR,
  FDR_threshold = 0.05,
)

sig_res_enrich_SR <- res_enrich_SR@result[res_enrich_SR@result$p.adjust < 0.05, ]
```

## LR-pathways

```{r}
DT::datatable(sig_res_enrich_LR,
              options = list(pageLength = 10, scrollX = TRUE),
              rownames = FALSE)

```

## SR-pathways

```{r}
DT::datatable(sig_res_enrich_SR,
              options = list(pageLength = 10, scrollX = TRUE),
              rownames = FALSE)
```

## Checking similar and different pathways between SR and LR

```{r}

gplots::venn(
  list(
   sig_res_enrich_SR$ID , sig_res_enrich_LR$ID)
  )


unique_GO_SR<- setdiff(sig_res_enrich_SR$Description, sig_res_enrich_LR$Description)

head(unique_GO_SR, 10)

unique_GO_LR<- setdiff(sig_res_enrich_LR$Description, sig_res_enrich_SR$Description)

head(unique_GO_LR, 10)


```

- 97 pathways were shared between SR and LR
- Both platforms detected increase in inflammatory pathways, SR detected more DDR pathways compared to LR but the overall picture is relatively similar

## Visualizing top enriched pathways

```{r}
barplot(res_enrich_LR, showCategory = 20)

barplot(res_enrich_SR, showCategory = 20)

```

# GSEA

```{r}

## background genes

background_genes <- results_LR_shrink%>%
  as.data.frame() %>%
  filter(baseMean != 0) %>%
  tibble::rownames_to_column(var = "gene") %>%
  pull(gene)


res_df <- results_LR_shrink %>%
  as.data.frame() %>%
  filter(baseMean != 0) %>%
  tibble::rownames_to_column(var = "gene")

background_genes_map <- bitr(
  geneID = background_genes,
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = "org.Hs.eg.db"
)


m_df <- msigdbr(species = "Homo sapiens")


m_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)


## rank all the genes by signed fold change * -log10pvalue.

res_df <- res_df %>%
  mutate(signed_rank_stats = sign(log2FoldChange) * -log10(pvalue)) %>%
  left_join(background_genes_map, by = c("gene" = "ENSEMBL")) %>%
  arrange(desc(signed_rank_stats))

gene_list <- res_df$signed_rank_stats
names(gene_list) <- res_df$ENTREZID

em2 <- GSEA(gene_list, TERM2GENE = m_t2g)

em2@result %>% View()

### visualization

p1 <- gseaplot(em2,
  geneSetID = "HALLMARK_DNA_REPAIR",
  by = "runningScore", title = "HALLMARK_DNA_REPAIR"
)

p2 <- gseaplot(em2,
  geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE",
  by = "runningScore", title = "HALLMARK_INFLAMMATORY_RESPONSE"
)

p1 / p2

```

## Creating GeneTonic object

```{r}

gtl_pilot_LR <- GeneTonicList(
  dds = dds_LR,
  res_de = results_LR,
  res_enrich = shake_enrichResult(res_enrich_LR),
  annotation_obj = anno_df
)

```

## GeneTonic Heatmap

```{r}

scored_gtl_mat <- GeneTonic::gs_scores(se = vsd_corrected_LR, gtl = gtl_pilot_LR)

anno_full <- as.data.frame(colData(vsd_phybion_LR)[, c("Sex", "Dose")])

column_annotation <- HeatmapAnnotation(
  Sex = anno_full$Sex,
  Dose = anno_full$Dose
)

GeneTonic::gs_scoresheat(scored_gtl_mat, n_gs = 20, cluster_cols = FALSE)


```


- Both SR and LR analysis yielded 115 DEGs, only 45 DEGs were shared between both.
- Although there was a huge difference between the DEGs, both platform yielded relatively similar picture at the pathway level.
- Out of the 70 DEGs that were unique to the LR, 46 gene were not detected as DEGs in SR full mode. 
- Among those 46 DEGs, 39 were captured by illumina but they were below significance (FDR 0.05)--> those 39 genes were mainly pseudogenes and ribosomal RNA
- out of the 34 novel genes detected in SR_full, only G0S2 was found in LR as DE gene and 5 gene in SR_4Gy (G0S2, NIBAN3, TIFA, PLEKHG1, MIR34AHG)
- 27181 gene_id were detected in dds_SR_4Gy, and 31525
- 27181 gene_id were detected in dds_SR_4Gy, and 31525--> 9852 gene_id that have rowsums > 0 were detected onlyin dds_LR 

# Session information

```{r sessioninfo}
sessionInfo()
```
