---
title: "PhyBioN_LR_Analysis"
date: "2025-11-27"
output: 
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_float: true
---
# Loading the libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library("DRIMSeq")
library("biomaRt")
library("ggplot2")
library("readxl")
library("readr")
library("BiocParallel")
library("tximport")
library("DESeq2")
library("edgeR")
library("mosdef")
library("limma")
library("stageR")
library("dplyr")
library("VennDiagram")
library("GenomicRanges")
library("rtracklayer")
library("ggrepel")
library("ggplot2")
library("ComplexHeatmap")
library("pheatmap")
library("GSVA")
library("msigdbr")
library("clusterProfiler")
library("org.Hs.eg.db")
library("enrichplot")
library("ReactomePA")
library("topGO")
library("quantiseqr")
library("circlize")
library("GeneTonic")
library("tibble")
library("Biostrings")
library("stringr")
```

# Loading the dataset

## Loading the long read -gene level- data
```{r}
dds_LR <- readRDS("~/Phybion_Pilot_LR/dds_LR.RDS")

# Variance stabilizing transformation

vsd_LR <- vst(dds_LR)
```

## Loading LR - Transcript level- data

```{r}
colData <- read_excel("./coldata_phybion_pilot_LR.xlsx")

colData <- as.data.frame(colData)


LR_metadata <- colData


LR_Tx_counts <- read.table(file = "./Isoquant/OUT/OUT.transcript_model_grouped_counts.tsv")

# Transcript counts
# Removing un-annotated transcripts since they cannot be used for diff splice analysis

LR_Tx_counts_filtered <- LR_Tx_counts[!grepl("^transcript", LR_Tx_counts$V1), ]

# Gencode to Ensemble id

LR_Tx_counts_filtered[, 1] <- gsub("\\..*", "", LR_Tx_counts_filtered[, 1])

# assigning the transcripts id to the rownames

rownames(LR_Tx_counts_filtered) <- LR_Tx_counts_filtered[, 1]
LR_Tx_counts_filtered <- LR_Tx_counts_filtered[, -1]


rownames(LR_Tx_counts) <- LR_Tx_counts[, 1]
LR_Tx_counts <- LR_Tx_counts[, -1]

LR_Tx_counts <- LR_Tx_counts[rowSums(LR_Tx_counts) > 6, ]
# naming the columns

colnames(LR_Tx_counts_filtered) <- c("0Gy-F", "0Gy-A", "0Gy-S", "4Gy-F", "4Gy-A", "4Gy-S")

colnames(LR_Tx_counts) <- c("0Gy-F", "0Gy-A", "0Gy-S", "4Gy-F", "4Gy-A", "4Gy-S")
```

# Data exploration

## PCA plot

```{r}
pca_plot_LR <- plotPCA(vsd_LR,
                       intgroup = c("Donors", "Dose"),
                       returnData = TRUE
)

percentVar <- round(100 * attr(pca_plot_LR, "percentVar"))



ggpt <- theme_bw() +
  theme(
    axis.title.x = element_text(size = rel(1.6)),
    axis.title.y = element_text(size = rel(1.6)),
    axis.text.x = element_text(size = rel(1.6)),
    axis.text.y = element_text(size = rel(1.6)),
    strip.text.x = element_text(size = rel(1.6)),
    strip.text.y = element_text(size = rel(1.6)),
    legend.text = element_text(size = rel(1.4)),
    legend.title = element_text(size = rel(1.4)),
    plot.title = element_text(size = rel(1.6))
  )


PCA_LR <- ggplot(
  pca_plot_LR,
  aes(x = PC1, y = PC2, color = Donors)
) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17)) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 17),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 13),
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 15)
  )

PCA_LR + ggpt

# There is a strong interdonor variation which has to be accounted for in the design before DEA
```

# EdgeR_DEA

## gene level DEA

```{r}
gene_mat <- counts(dds_LR)

LR_samples <- data.frame(
  sample_id = LR_metadata$sample_id,
  group = LR_metadata$Dose,
  Donors = LR_metadata$Donors
)

knitr::kable(LR_samples)

# Create DGEList
y_DEA <- DGEList(
  counts = gene_mat,
  samples = LR_samples
)

keep <- filterByExpr(y_DEA)
y_DEA <- y_DEA[keep, keep.lib.sizes = FALSE]

# TMM normalization
y_DEA <- normLibSizes(y_DEA)

# Define factors
Donor <- factor(LR_metadata$Donors)
Treatment <- factor(LR_metadata$Dose)

# Design matrix including donor
design <- model.matrix(~ Donor + Treatment)

#  Estimate dispersions

y_DEA <- estimateDisp(y_DEA, design)

y_DEA$common.dispersion

# dispersion estimate plot

plotBCV(y_DEA)

#  Fit GLM model
fit_gene <- glmQLFit(y_DEA, design)

# plot quasi-likelihood despersion

plotQLDisp(fit_gene)

# Test for a treatment effect
lrt_gene <- glmQLFTest(fit_gene, coef = "Treatment4Gy")

# Extract results
edger_gene_res <- topTags(lrt_gene, n = Inf)


sig_genes_edger <- edger_gene_res$table[edger_gene_res$table$FDR <= 0.05, ]

# plot raw pvalues histogram

hist(edger_gene_res$table$PValue)


sig_genes_edger$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = rownames(sig_genes_edger),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```

## Volcanoplot

```{r}
de_res <- edger_gene_res$table

top_genes <- de_res[order(de_res$FDR), ][1:10, ]
top_genes$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = rownames(top_genes),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)


# Define up/down/NS
de_res$diffexpressed <- factor(ifelse(de_res$FDR < 0.05 & de_res$logFC > 0, "Up",
                                      ifelse(de_res$FDR < 0.05 & de_res$logFC < 0, "Down", "NS")
))

# Volcano plot
g <- ggplot(de_res, aes(x = logFC, y = -log10(FDR), color = diffexpressed)) +
  geom_point(alpha = 0.4, size = 3) +
  geom_text_repel(
    data = top_genes,
    aes(x = logFC, y = -log10(FDR), label = SYMBOL),
    size = 4,
    color = "black"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  xlim(c(-14, 14)) +
  xlab("log2 fold change") +
  ylab("-log10 FDR") +
  scale_color_manual(values = c("blue", "grey", "red")) +
  theme_bw() +
  theme(legend.position = "none") +
  ggpt

g
```
## Creating gene plotting function

```{r}
plot_gene_modified <- function(dds, gene, intgroup = "condition", assay = "counts",
                               annotation_obj = NULL, normalized = TRUE, transform = TRUE,
                               labels_display = FALSE, labels_repel = TRUE, plot_type = "auto",
                               return_data = FALSE, color_by = "Donors") {
  plot_type <- match.arg(plot_type, c("auto", "jitteronly", "boxplot", "violin", "sina"))
  
  if (!(all(intgroup %in% colnames(colData(dds))))) {
    stop("`intgroup` not found in the colData slot of the dds object.")
  }
  
  df <- get_expr_values(
    de_container = dds, gene = gene, intgroup = intgroup,
    assay = assay, normalized = normalized
  )
  df$sample_id <- rownames(df)
  
  # Add Donors column if it exists
  if ("Donors" %in% colnames(colData(dds))) {
    df$Donors <- colData(dds)[, "Donors"]
  } else {
    stop("The `Donors` column is missing in the colData of the dds object.")
  }
  
  # Add gene annotation if provided
  if (!is.null(annotation_obj)) {
    genesymbol <- annotation_obj$gene_name[match(gene, annotation_obj$gene_id)]
  } else {
    genesymbol <- gene
  }
  
  onlyfactors <- df[, match(intgroup, colnames(df))]
  df$plotby <- interaction(onlyfactors)
  min_by_groups <- min(table(df$plotby))
  
  if (return_data) {
    return(df)
  }
  
  p <- ggplot(df, aes(x = .data$plotby, y = .data$exp_value, col = .data[[color_by]])) +
    scale_x_discrete(name = "") +
    scale_color_discrete(name = "Experimental\ngroup") +
    theme_bw()
  
  jit_pos <- position_jitter(width = 0.2, height = 0, seed = 42)
  
  # Plot individual points
  p <- p + geom_point(position = jit_pos, shape = 16, size = 3)
  
  # Display labels if required
  if (labels_display) {
    if (labels_repel) {
      p <- p + ggrepel::geom_text_repel(aes(label = .data$sample_id),
                                        min.segment.length = 0, position = jit_pos
      )
    } else {
      p <- p + geom_text(aes(label = .data$sample_id), hjust = -0.1, vjust = 0.1, position = jit_pos)
    }
  }
  
  # Add a connected mean line across all groups
  #  p <- p + stat_summary(fun = mean, geom = "line", aes(group = 1),
  #                       color = "grey80", linewidth = 0.8)
  
  y_label <- if (assay == "counts" & normalized) {
    "Normalized counts"
  } else if (assay == "counts" & !normalized) {
    "Counts"
  } else if (assay == "abundance") {
    "TPM - Transcripts Per Million"
  } else {
    assay
  }
  
  if (transform) {
    p <- p + scale_y_log10(name = paste0(y_label, " (log10 scale)"))
  } else {
    p <- p + scale_y_continuous(name = y_label)
  }
  
  p <- p + labs(title = paste0(genesymbol, " - ", gene))
  
  return(p)
}
```


## Checking the gene pattern across doses

```{r}
anno_df <- get_annotation_orgdb(
  de_container = dds_LR,
  orgdb_package = "org.Hs.eg.db",
  id_type = "ENSEMBL"
)


gene_plot <- plot_gene_modified(
  dds = dds_LR,
  gene = c("ENSG00000124762"),
  intgroup = "Dose",
  annotation_obj = anno_df
) + ggpt

gene_plot

# ggsave("cd83.png", plot = gene_plot, width = 8, height = 5, dpi = 400)
```

# plotting tx

```{r}
LR_Tx_counts_full <- read.table(file = "./Isoquant/OUT/OUT.transcript_model_grouped_counts.tsv")

tx_mat <- LR_Tx_counts_full
rownames(tx_mat) <- tx_mat$V1
tx_mat <- tx_mat[, -1]

colnames(tx_mat) <- c("0Gy-F", "0Gy-A", "0Gy-S", "4Gy-F", "4Gy-A", "4Gy-S")


dds_tx <- DESeqDataSetFromMatrix(
  countData = tx_mat,
  colData = LR_metadata,
  design = ~ Dose + Donors
)

# ENST00000440815 ENST00000325568

plot_gene_modified(
  dds = dds_tx,
  gene = c("ENST00000325568.9"),
  intgroup = "Dose",
  annotation_obj = anno_df
) +
  labs(
    title = "IL-32 - ENST00000325568"
  ) + ggpt
```

## transcript level DEA

```{r}
tx_mat <- LR_Tx_counts

y_DEA_tx <- DGEList(
  counts = tx_mat,
  samples = LR_samples
)

keep <- filterByExpr(y_DEA_tx)
y_DEA_tx <- y_DEA_tx[keep, , keep.lib.sizes = FALSE]

# TMM normalization
y_DEA_tx <- normLibSizes(y_DEA_tx)

# Define factors
Donor <- factor(LR_metadata$Donors)
Treatment <- factor(LR_metadata$Dose)

# Design matrix including donor
design <- model.matrix(~ Donor + Treatment)

#  Estimate dispersions
y_DEA_tx <- estimateDisp(y_DEA_tx, design)

y_DEA_tx$common.dispersion

# dispersion estimate plot

plotBCV(y_DEA_tx)

# Fit GLM model
fit_tx <- glmQLFit(y_DEA_tx, design)

# plot quasi-likelihood despersion

plotQLDisp(fit_tx)

# Test for a treatment effect

lrt_tx <- glmQLFTest(fit_tx, coef = "Treatment4Gy")

# Extract results
edger_Tx_res <- topTags(lrt_tx, n = Inf)


sig_tx <- edger_Tx_res$table[edger_Tx_res$table$FDR <= 0.05, ]

# Filter rows starting with "transcript"
sig_novel_tx <- sig_tx[grepl("^transcript", rownames(sig_tx)), ]

# plot raw pvalues histogram

hist(edger_Tx_res$table$PValue)
```

# diffsplice- DTU

## Preparing the matrix for the analysis

```{r}
LR_Tx_counts_filtered <- as.data.frame(LR_Tx_counts_filtered)
LR_Tx_counts_filtered$feature_id <- rownames(LR_Tx_counts_filtered)

# extracting and merging gene_id to LR_counts

mart <- useEnsembl(
  biomart = "ensembl",
  dataset = "hsapiens_gene_ensembl"
)


transcript_to_gene <- getBM(
  attributes = c("ensembl_transcript_id", "external_gene_name"),
  filters = "ensembl_transcript_id",
  values = LR_Tx_counts_filtered$feature_id,
  mart = mart
)


LR_Tx_counts_filtered <- merge(LR_Tx_counts_filtered, transcript_to_gene,
                               by.x = "feature_id",
                               by.y = "ensembl_transcript_id",
                               all.x = TRUE
)

# naming and reordering

colnames(LR_Tx_counts_filtered)[8] <- "gene_id"

LR_Tx_counts_filtered <- LR_Tx_counts_filtered[, c(1, 8, 2, 3, 4, 5, 6, 7)]

LR_Tx_counts_filtered <- na.omit(LR_Tx_counts_filtered)
```

## running diffsplice
```{r}
mat <- as.matrix(LR_Tx_counts_filtered[, -c(1, 2)])
rownames(mat) <- LR_Tx_counts_filtered$feature_id

gene_df <- LR_Tx_counts_filtered[, c(1, 2)]


y <- DGEList(counts = mat, samples = LR_samples, genes = gene_df)

keep <- filterByExpr(y, group = LR_metadata$Dose)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- normLibSizes(y)
y$samples

design <- model.matrix(~ Donors + group, data = y$samples)

fit <- glmQLFit(y, design)

ex <- diffSplice(
  fit = fit,
  coef = 4,
  geneid = "gene_id",
  exonid = "feature_id",
  robust = TRUE
)

ts <- topSplice(ex, number = Inf)
ts.tx <- topSplice(ex, test = "t", number = Inf)
```

## StageR

```{r}
pScreen <- ts$P.Value
names(pScreen) <- ts$gene_id
pConfirmation <- matrix(ts.tx$P.Value, ncol = 1)
rownames(pConfirmation) <- ts.tx$feature_id

tx2gene <- ts.tx[, c(1, 2)]

stageRObj2 <- stageRTx(
  pScreen = pScreen,
  pConfirmation = pConfirmation,
  pScreenAdjusted = FALSE,
  tx2gene = tx2gene
)

stageRObj2 <- stageWiseAdjustment(
  object = stageRObj2,
  method = "dtu",
  alpha = 0.05
)
suppressWarnings({
  ds.padj <- getAdjustedPValues(stageRObj2,
                                order = TRUE,
                                onlySignificantGenes = TRUE
  )
})
head(ds.padj)

ds.filtered <- ds.padj[ds.padj$transcript <= 0.05, ]


plotSplice(ex, geneid = "RPS19", genecolname = "gene_id", FDR = 0.05)
```

# Enrichment --> Kegg, GO, reactome

```{r}
deg_genes <- rownames(sig_genes_edger)

# Convert gene symbols to Entrez IDs
entrez_genes <- bitr(deg_genes,
                     fromType = "ENSEMBL",
                     toType = "ENTREZID",
                     OrgDb = org.Hs.eg.db
)

# Perform KEGG pathway enrichment
kegg_res <- enrichKEGG(
  gene = entrez_genes$ENTREZID,
  organism = "hsa", # "hsa" for human
  pvalueCutoff = 0.05
)

kegg_res <- kegg_res@result[kegg_res@result$p.adjust <= 0.05, ]


# reactome pathway analysis

reactome_res <- enrichPathway(
  gene = entrez_genes$ENTREZID,
  organism = "human",
  pvalueCutoff = 0.05,
  readable = TRUE
)

reactome_res <- reactome_res@result[reactome_res@result$p.adjust <= 0.05, ]

# GO enrichment

deg_genes_symbol <- mapIds(
  org.Hs.eg.db,
  keys = deg_genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

clusterprofiler_res <- enrichGO(deg_genes_symbol,
                                keyType = "SYMBOL",
                                ont = "BP",
                                OrgDb = "org.Hs.eg.db",
                                pvalueCutoff = 0.05,
                                pAdjustMethod = "BH",
                                qvalueCutoff = 0.2,
                                minGSSize = 10,
                                maxGSSize = 500
)


clusterprofiler_res_sig <- clusterprofiler_res@result[clusterprofiler_res@result$p.adjust <= 0.05, ]

View(clusterprofiler_res_sig)
```

## GeneTonic

```{r}
# create DESeq result object

de_res_table <- edger_gene_res$table

deseq_like <- DataFrame(
  log2FoldChange = de_res_table$logFC,
  stat = de_res_table$F,
  pvalue = de_res_table$PValue,
  padj = de_res_table$FDR,
  row.names = rownames(de_res_table)
)

deseq_like <- DESeqResults(deseq_like)

# create GeneTonic object

gtl_4Gy <- GeneTonicList(
  dds = dds_LR,
  res_de = deseq_like,
  res_enrich = shake_enrichResult(clusterprofiler_res),
  annotation_obj = anno_df
)

# GeneTonic(gtl= gtl_4Gy)


# preparing the matrix for heatmap

vsd_LR_corrected <- vsd_LR

assay(vsd_LR_corrected) <- limma::removeBatchEffect(assay(vsd_LR_corrected), batch = vsd_LR_corrected$Donors)
scored_mat <- GeneTonic::gs_scores(se = vsd_LR_corrected, gtl = gtl_4Gy)


anno <- as.data.frame(colData(vsd_LR_corrected)[, c("Sex", "Donors", "Dose")])

dose_colors <- colorRampPalette(c("lightyellow", "red"))(2)
names(dose_colors) <- c("0Gy", "4Gy")

anno_colors <- list(
  Sex = c(female = "pink", male = "blue"),
  Dose = dose_colors
)


column_annotation <- HeatmapAnnotation(
  Sex = anno$Sex,
  Dose = anno$Dose,
  col = anno_colors, show_legend = T
)

library(circlize)

col_fun <- colorRamp2(
  c(-2, 0, 2),
  c("blue", "white", "red")
)

ht <- Heatmap(scored_mat[1:15, ],
              name = "Expression",
              top_annotation = column_annotation,
              cluster_columns = F, # Optional: Cluster columns
              cluster_rows = F, # Optional: Cluster rows
              show_row_names = T,
              show_column_names = TRUE,
              row_dend_side = "left",
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 10, fontface = "bold"),
              width = unit(7, "cm"),
              heatmap_height = unit(12, "cm"),
              column_names_gp = gpar(fontsize = 7),
              column_labels = rep(c(1, 2, 3), 2),
              column_names_rot = 360,
              column_title = "Donors",
              column_title_side = "bottom",
              show_heatmap_legend = T,
              col = col_fun
)


draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right")
```

## enrichment for the downregulated genes

```{r}
dwn_sig <- sig_genes_edger[sig_genes_edger$logFC < 0, ]

dwn_genes <- dwn_sig$SYMBOL

clusterprofiler_res_dwn <- enrichGO(dwn_genes,
                                    keyType = "SYMBOL",
                                    ont = "BP",
                                    OrgDb = "org.Hs.eg.db",
                                    pvalueCutoff = 0.05,
                                    pAdjustMethod = "BH",
                                    qvalueCutoff = 0.2,
                                    minGSSize = 10,
                                    maxGSSize = 500
)


clusterprofiler_res_dwn <- clusterprofiler_res_dwn@result[clusterprofiler_res_dwn@result$p.adjust <= 0.05, ]

View(clusterprofiler_res_dwn)
```

# Gene Set Variation Analysis-GSVA

```{r}
exprs_mat <- assay(dds_LR)

exprs_mat <- exprs_mat[!grepl("^transcript", rownames(exprs_mat)), ]


exprs_mat <- exprs_mat[rowSums(exprs_mat) > 6, ]


genes_exprs_mat <- rownames(exprs_mat)

genes_exprs_mat <- mapIds(
  org.Hs.eg.db,
  keys = genes_exprs_mat,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

exprs_df <- as.data.frame(exprs_mat)
exprs_df$SYMBOL <- genes_exprs_mat

# calculating the variance

exprs_df$variance <- apply(exprs_df[, !(names(exprs_df) %in% c("SYMBOL"))], 1, var, na.rm = TRUE)

# 2. Keep the most variable transcript per gene symbol
exprs_highvar <- exprs_df %>%
  group_by(SYMBOL) %>%
  slice_max(order_by = variance, n = 1, with_ties = FALSE) %>%
  ungroup()

# 3. Remove helper column
exprs_highvar <- exprs_highvar %>% dplyr::select(-variance)

# 4. Set SYMBOL as rownames
exprs_highvar <- as.data.frame(exprs_highvar)

exprs_highvar <- na.omit(exprs_highvar)
rownames(exprs_highvar) <- exprs_highvar$SYMBOL
exprs_highvar$SYMBOL <- NULL

exprs_highvar <- as.matrix(exprs_highvar)

# run GSVA

goannot <- AnnotationDbi::select(org.Hs.eg.db, keys = keys(org.Hs.eg.db, keytype = "SYMBOL"), columns = "GOALL", keytype = "SYMBOL")

genesbygo <- split(goannot$SYMBOL, goannot$GO)

huangPar <- gsvaParam(exprs_highvar, genesbygo, minSize = 5, maxSize = 2000)

esmicro <- gsva(huangPar)

# adding a column for GO id description

go_ids <- rownames(esmicro)

# Get the term names
go_terms <- Term(GOTERM[go_ids])
# Add as a new column
esmicro <- as.data.frame(esmicro)
esmicro$Description <- go_terms
rownames(esmicro) <- esmicro$Description
esmicro <- esmicro[, -7]


# Applying Limma on GSVA matrix to extract differentially expressed pathways

# creating metadata

patient <- rep(c("p1", "p2", "p3"), 2)
dose <- c(rep("0Gy", 3), rep("4Gy", 3))
metadata <- data.frame(patient, dose)

# make the design

f <- factor(paste(metadata$dose))

design <- model.matrix(~ 0 + f)

patient <- as.factor(metadata$patient)

# limma-based DEA

corfit <- duplicateCorrelation(esmicro, design, block = patient)

fit <- lmFit(esmicro, design, block = patient, correlation = corfit$consensus)

contrast.matrix <- makeContrasts((f4Gy - f0Gy), levels = design)

fit2 <- contrasts.fit(fit, contrast.matrix)

fit2 <- eBayes(fit2)

results_gsva <- topTable(fit2, coef = 1, number = nrow(esmicro), adjust = "BH")

results_f_gsva <- decideTests(fit2, p.value = 0.05)

summary(results_f_gsva)


results_f_gsva <- as.data.frame(results_f_gsva)


sig_gsva <- results_f_gsva[results_f_gsva[, 1] == 1, ]
```

## GSEA using GO terms

```{r}
## background genes

background_genes <- edger_gene_res$table %>%
  as.data.frame() %>%
  filter(logCPM != 0) %>%
  tibble::rownames_to_column(var = "gene") %>%
  pull(gene)


res_df <- edger_gene_res$table %>%
  as.data.frame() %>%
  filter(logCPM != 0) %>%
  tibble::rownames_to_column(var = "gene")

background_genes_map <- bitr(
  geneID = background_genes,
  fromType = "ENSEMBL",
  toType = "SYMBOL",
  OrgDb = "org.Hs.eg.db"
)


m_df <- msigdbr(species = "Homo sapiens")

head(m_df)

go_df <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% dplyr::select(gs_name, gene_symbol)



# table(m_t2g$gs_name)
# head(m_t2g)

### Gene set enrichment analysis

## rank all the genes by signed fold change * -log10pvalue.

res_df <- res_df %>%
  mutate(signed_rank_stats = sign(logFC) * -log10(PValue)) %>%
  left_join(background_genes_map, by = c("gene" = "ENSEMBL")) %>%
  arrange(desc(signed_rank_stats))

# removing duplicates by choosing the ones with FDR <=0.05

res_df_filtered <- res_df %>%
  dplyr::group_by(SYMBOL) %>%
  dplyr::arrange(FDR, .by_group = TRUE) %>%
  dplyr::slice( # keep logic per group
    if (any(FDR <= 0.05)) which.min(FDR) else 1
  ) %>%
  dplyr::ungroup()


res_df_filtered <- na.omit(res_df_filtered)

gene_list <- res_df_filtered$signed_rank_stats
names(gene_list) <- res_df_filtered$SYMBOL
gene_list <- sort(gene_list, decreasing = TRUE)


em2 <- GSEA(gene_list, TERM2GENE = go_df)

em2@result %>% View()

write.csv(em2@result, "GSEA_results.csv")

### visualization

p1 <- gseaplot(em2,
               geneSetID = "GOBP_PEPTIDE_ANTIGEN_ASSEMBLY_WITH_MHC_CLASS_II_PROTEIN_COMPLEX",
               by = "runningScore", title = "GOBP_PEPTIDE_ANTIGEN_ASSEMBLY_WITH_MHC_CLASS_II_PROTEIN_COMPLEX"
)

p2 <- gseaplot(em2,
               geneSetID = "GOBP_HUMORAL_IMMUNE_RESPONSE",
               by = "runningScore", title = "GOBP_HUMORAL_IMMUNE_RESPONSE"
)

p1 / p2

# ggsave("GSEA.png", plot = p1 / p2, width = 7, height = 6, dpi = 400)
```



# Transdecoder

```{r}
# Read protein FASTA file
pep <- readAAStringSet("novel_isoforms_n.fa.transdecoder.pep")

# Inspect
pep
names(pep)[1:5] # transcript IDs
width(pep)[1:5] # sequence lengths
pep[1] # first sequence

# convert to df
headers <- names(pep)
seqs <- as.character(pep)

# Parse each header
pep_df <- tibble(
  transcript_id = headers,
  sequence = seqs,
  orf_type = str_extract(headers, "(?<=ORF type:)\\w+"),
  strand = str_extract(headers, "(?<=\\()(\\+|\\-)\\)(?=,score=)"),
  score = str_extract(headers, "(?<=score=)[0-9.]+") %>% as.numeric(),
  length = str_extract(headers, "(?<=len:)\\d+") %>% as.numeric(),
  start = str_extract(headers, "(?<=:)\\d+(?=-)") %>% as.numeric(),
  end = str_extract(headers, "(?<=-)\\d+(?=\\([+-]\\))") %>% as.numeric(),
  coord_strand = str_extract(headers, "(?<=\\d\\()[+-](?=\\))")
)

pep_df <- pep_df %>%
  mutate(final_strand = ifelse(!is.na(coord_strand), coord_strand, strand)) %>%
  dplyr::select(-strand, -coord_strand)

# filter on the tx that are potentially coding genes

pep_df_strict <- pep_df %>%
  dplyr::filter(
    score >= 100,
    orf_type == "complete"
  )


pep_df$clean_id <- sub("^([^.]+\\.[^.]+\\.[^.]+).*", "\\1", pep_df[[1]])

pep_df_strict$clean_id <- sub("^([^.]+\\.[^.]+\\.[^.]+).*", "\\1", pep_df_strict[[1]])
```

# CPC2

```{r}
novel_isoforms_fa <- readAAStringSet("novel_isoforms_n.fa")



novel_isoforms_fa_df <- as.data.frame(novel_isoforms_fa)

# focusing on the 8 transcripts detected by transdecoder

selected_8_tx <- c("transcript13458.chr22.nnic", "transcript13760.chr9.nnic", "transcript13826.chr15", "transcript15213.chr10.nnic", "transcript21609.chr6.nnic", "transcript21664.chr9.nnic", "transcript23187.chr5.nnic", "transcript36523.chr12.nnic")

selected_tx_fa <- novel_isoforms_fa_df[selected_8_tx, ]


ids <- paste0("seq", seq_along(selected_tx_fa))

fa_file <- "selected_tx.fa"
writeLines(
  paste0(">", ids, "\n", selected_tx_fa),
  con = fa_file
)



cpc_res <- read.table("cpc_res.txt")
colnames(cpc_res) <- c("ID", "peptide_length", "Fickett_score", "pI", "ORF_integrity", "coding_probability", "label")

cpc_res_coding <- cpc_res[cpc_res$label == "coding", ]
cpc_res_noncoding <- cpc_res[cpc_res$label == "noncoding", ]
```

# Gffcompare

```{r}
tmap <- read.table("Isoquant/OUT/gffcmp_out.OUT.transcript_models.gtf.tmap",
                   sep = "\t", header = FALSE, quote = "", stringsAsFactors = FALSE, fill = TRUE
)

# 3. Filter the tmap file to keep only those novel transcripts

selected_8_tx <- c("transcript13458.chr22.nnic", "transcript13760.chr9.nnic", "transcript13826.chr15.nnic", "transcript15213.chr10.nnic", "transcript21609.chr6.nnic", "transcript21664.chr9.nnic", "transcript23187.chr5.nnic", "transcript36523.chr12.nnic")


tmap_novel <- tmap %>%
  filter(V5 %in% selected_8_tx)

tmap_novel_full <- tmap %>%
  filter(V5 %in% rownames(sig_novel_tx))

colnames(tmap_novel) <- tmap[1, ]
colnames(tmap_novel_full) <- tmap[1, ]

tmap_novel_full[, 1] <- gsub("\\..*", "", tmap_novel_full[, 1])

tmap_novel_full$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = tmap_novel_full$ref_gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```

# CPAT

```{r}
cpat_res <- read_tsv("cpat_results.txt.ORF_prob.best.tsv")

cpat_res_coding <- cpat_res[cpat_res$Coding_prob >= 0.5, ]

View(cpat_res_coding)
```

```{r}
venn.plot <- venn.diagram(
  x = list(
    TransDecoder = unique(pep_df$clean_id),
    CPC2 = cpc_res_coding$ID,
    CPAT = tolower(cpat_res_coding$seq_ID)
  ),
  filename = NULL, # Plot to R directly
  fill = c("#66c2a5", "#8da0cb", "orange"),
  alpha = 0.6,
  cat.cex = 1.2,
  cex = 1.5,
  #  main = "Novel coding isoforms"
)

grid::grid.draw(venn.plot)


# the transcript that exist only on cpat (transcript37812.chr6.nnic), if at all, could be a coding one. it is labeled non coding in cpc 2 but its coding probability is 0.34. however, it doesnt exist at all in transdecoder output
# the transcript that exist in transdecoder only (transcript25470.chr1.nic) doesnt exist at all in cpat output and it as very low coding prob in cpc2 0.00349659


# out of the 2 transcripts that were shared between transdecoder and cpat but not cpc2 is CDKN1a. this one has high potentiality for being coding as well. CPC score is 0.44, gffcompare class code is j, cpat score is 0.8. ORF length is 120 aa, has 3 exons, nucleotide seq of 1170, and complete ORF type. also 26% query coverage with blastx


TransDecoder <- tolower(unique(pep_df$clean_id))
CPC2 <- tolower(cpc_res_coding$ID)
CPAT <- tolower(cpat_res_coding$seq_ID)

shared_transcripts <- Reduce(intersect, list(TransDecoder, CPC2, CPAT))
length(shared_transcripts)
```


```{r}
# For CPAT: classify based on coding probability threshold
cpat_res_labeled <- cpat_res %>%
  mutate(
    tool = "CPAT",
    label = ifelse(Coding_prob >= 0.5, "Coding", "Non-coding")
  )

# For CPC2: assume the column is named 'coding' or similar
cpc_res_labeled <- cpc_res %>%
  mutate(
    tool = "CPC2",
    label = ifelse(tolower(label) == "coding", "Coding", "Non-coding")
  )

# --- TransDecoder ---
# You mentioned there are 14 total transcripts (11 unique) with ORFs
# Let's get unique ones
transdecoder_unique <- unique(pep_df$clean_id)

# Total number of isoforms in your dataset (e.g., 26 total)
total_isoforms <- 26 # <-- change this to your real total number

# Compute counts
transdecoder_df <- tibble(
  tool = "TransDecoder",
  label = c("Coding", "Non-coding"),
  count = c(length(transdecoder_unique), total_isoforms - length(transdecoder_unique))
)


# Count per tool and label
summary_df <- bind_rows(cpat_res_labeled, cpc_res_labeled) %>%
  count(tool, label, name = "count")

# Add TransDecoder data
summary_df <- bind_rows(summary_df, transdecoder_df)

ggplot(summary_df, aes(x = tool, y = count, fill = label)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Coding" = "steelblue", "Non-coding" = "orange")) +
  labs(
    title = "Comparison of Coding vs Non-coding Transcripts Across Tools",
    x = "Tool",
    y = "Count",
    fill = "Transcript Type"
  ) +
  theme_minimal(base_size = 14)
```

# Session info

```{r}
sessionInfo()
```
