---
title: "PhyBioN_LR_Analysis"
date: "2025-11-27"
output: 
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_float: true
---
# Loading the libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library("DRIMSeq")
library("biomaRt")
library("ggplot2")
library("readxl")
library("readr")
library("BiocParallel")
library("tximport")
library("DESeq2")
library("edgeR")
library("mosdef")
library("limma")
library("stageR")
library("dplyr")
library("VennDiagram")
library("GenomicRanges")
library("rtracklayer")
library("ggrepel")
library("ggplot2")
library("ComplexHeatmap")
library("pheatmap")
library("GSVA")
library("msigdbr")
library("clusterProfiler")
library("org.Hs.eg.db")
library("enrichplot")
library("ReactomePA")
library("topGO")
library("quantiseqr")
library("circlize")
library("GeneTonic")
library("tibble")
library("Biostrings")
library("stringr")
```

# Loading the dataset

## Loading the long read -gene level- data
```{r}
dds_LR <- readRDS("dds_LR_pseudonymized.RDS")

# Variance stabilizing transformation

vsd_LR <- vst(dds_LR)
```

## Loading LR - Transcript level- data

```{r}
colData <- read_excel("./coldata_phybion_pilot_LR.xlsx")

colData <- as.data.frame(colData)


LR_metadata <- colData


LR_Tx_counts <- read.table(file = "./Isoquant/OUT/OUT.transcript_model_grouped_counts.tsv")

# Transcript counts
# Removing un-annotated transcripts since they cannot be used for diffsplice analysis

LR_Tx_counts_filtered <- LR_Tx_counts[!grepl("^transcript", LR_Tx_counts$V1), ]

# Gencode to Ensemble id

LR_Tx_counts_filtered[, 1] <- gsub("\\..*", "", LR_Tx_counts_filtered[, 1])

# assigning the transcripts id to the rownames

rownames(LR_Tx_counts_filtered) <- LR_Tx_counts_filtered[, 1]
LR_Tx_counts_filtered <- LR_Tx_counts_filtered[, -1]


rownames(LR_Tx_counts) <- LR_Tx_counts[, 1]
LR_Tx_counts <- LR_Tx_counts[, -1]

LR_Tx_counts <- LR_Tx_counts[rowSums(LR_Tx_counts) > 6, ]
# naming the columns

colnames(LR_Tx_counts_filtered) <- c("0Gy-do1", "0Gy-do2", "0Gy-do3", "4Gy-do1", "4Gy-do2", "4Gy-do3")

colnames(LR_Tx_counts) <- c("0Gy-do1", "0Gy-do2", "0Gy-do3", "4Gy-do1", "4Gy-do2", "4Gy-do3")
```

# Data exploration

## PCA plot

```{r}

pca_plot_LR <- plotPCA(vsd_LR,
                       intgroup = c("Donors", "Dose"),
                       returnData = TRUE
)

percentVar <- round(100 * attr(pca_plot_LR, "percentVar"))



ggpt <- theme_bw() +
  theme(
    axis.title.x = element_text(size = rel(1.6)),
    axis.title.y = element_text(size = rel(1.6)),
    axis.text.x = element_text(size = rel(1.6)),
    axis.text.y = element_text(size = rel(1.6)),
    strip.text.x = element_text(size = rel(1.6)),
    strip.text.y = element_text(size = rel(1.6)),
    legend.text = element_text(size = rel(1.4)),
    legend.title = element_text(size = rel(1.4)),
    plot.title = element_text(size = rel(1.6))
  )


PCA_LR <- ggplot(
  pca_plot_LR,
  aes(x = PC1, y = PC2, color = Donors)
) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17)) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 17),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 13),
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 15)
  )

PCA_LR + ggpt

# There is a strong interdonor variation which has to be accounted for in the design before DEA
```

## Creating gene plotting function

```{r}
plot_gene_modified <- function(dds, gene, intgroup = "condition", assay = "counts",
                               annotation_obj = NULL, normalized = TRUE, transform = TRUE,
                               labels_display = FALSE, labels_repel = TRUE, plot_type = "auto",
                               return_data = FALSE, color_by = "Donors") {
  plot_type <- match.arg(plot_type, c("auto", "jitteronly", "boxplot", "violin", "sina"))
  
  if (!(all(intgroup %in% colnames(colData(dds))))) {
    stop("`intgroup` not found in the colData slot of the dds object.")
  }
  
  df <- get_expr_values(
    de_container = dds, gene = gene, intgroup = intgroup,
    assay = assay, normalized = normalized
  )
  df$sample_id <- rownames(df)
  
  # Add Donors column if it exists
  if ("Donors" %in% colnames(colData(dds))) {
    df$Donors <- colData(dds)[, "Donors"]
  } else {
    stop("The `Donors` column is missing in the colData of the dds object.")
  }
  
  # Add gene annotation if provided
  if (!is.null(annotation_obj)) {
    genesymbol <- annotation_obj$gene_name[match(gene, annotation_obj$gene_id)]
  } else {
    genesymbol <- gene
  }
  
  onlyfactors <- df[, match(intgroup, colnames(df))]
  df$plotby <- interaction(onlyfactors)
  min_by_groups <- min(table(df$plotby))
  
  if (return_data) {
    return(df)
  }
  
  p <- ggplot(df, aes(x = .data$plotby, y = .data$exp_value, col = .data[[color_by]])) +
    scale_x_discrete(name = "") +
    scale_color_discrete(name = "Experimental\ngroup") +
    theme_bw()
  
  jit_pos <- position_jitter(width = 0.2, height = 0, seed = 42)
  
  # Plot individual points
  p <- p + geom_point(position = jit_pos, shape = 16, size = 4)
  
  # Display labels if required
  if (labels_display) {
    if (labels_repel) {
      p <- p + ggrepel::geom_text_repel(aes(label = .data$sample_id),
                                        min.segment.length = 0, position = jit_pos
      )
    } else {
      p <- p + geom_text(aes(label = .data$sample_id), hjust = -0.1, vjust = 0.1, position = jit_pos)
    }
  }
  
  
  y_label <- if (assay == "counts" & normalized) {
    "Normalized counts"
  } else if (assay == "counts" & !normalized) {
    "Counts"
  } else if (assay == "abundance") {
    "TPM - Transcripts Per Million"
  } else {
    assay
  }
  
  if (transform) {
    p <- p + scale_y_log10(name = paste0(y_label, " (log10 scale)"))
  } else {
    p <- p + scale_y_continuous(name = y_label)
  }
  
  p <- p + labs(title = paste0(genesymbol, " - ", gene))
  
  return(p)
}
```


```{r}
plot_gene_modified <- function(dds, gene, intgroup = "condition", assay = "counts",
                               annotation_obj = NULL, normalized = TRUE, transform = TRUE,
                               labels_display = FALSE, labels_repel = TRUE, plot_type = "auto",
                               return_data = FALSE,
                               color_by = "Dose",
                               shape_by = "Donors",
                               dose_colors = NULL) {
    
    plot_type <- match.arg(plot_type, c("auto", "jitteronly", "boxplot", "violin", "sina"))
    
    if (!(all(intgroup %in% colnames(colData(dds))))) {
        stop("`intgroup` not found in the colData slot of the dds object.")
    }
    
    df <- get_expr_values(
        de_container = dds, gene = gene, intgroup = intgroup,
        assay = assay, normalized = normalized
    )
    df$sample_id <- rownames(df)
    
    ## Add Donors
    if ("Donors" %in% colnames(colData(dds))) {
        df$Donors <- colData(dds)[, "Donors"]
    } else {
        stop("The `Donors` column is missing in the colData of the dds object.")
    }
    
    ## Gene annotation
    genesymbol <- if (!is.null(annotation_obj)) {
        annotation_obj$gene_name[match(gene, annotation_obj$gene_id)]
    } else {
        gene
    }
    
    onlyfactors <- df[, match(intgroup, colnames(df))]
    df$plotby <- interaction(onlyfactors)
    
    if (return_data) return(df)
    
    ## ---- PLOT ----
    p <- ggplot(
        df,
        aes(
            x = .data$plotby,
            y = .data$exp_value,
            color = .data[[color_by]],
            shape = .data[[shape_by]]
        )
    ) +
        scale_x_discrete(name = "") +
        theme_bw()
    
    ## Manual colors for doses (if provided)
    if (!is.null(dose_colors)) {
        p <- p + scale_color_manual(
            values = dose_colors,
            name = color_by
        )
    } else {
        p <- p + scale_color_discrete(name = color_by)
    }
    
    ## Shapes (donors)
    p <- p + scale_shape_discrete(name = shape_by)
    
    jit_pos <- position_jitter(width = 0.2, height = 0, seed = 42)
    
    p <- p + geom_point(position = jit_pos, size = 4)
    
    ## Labels
    if (labels_display) {
        if (labels_repel) {
            p <- p + ggrepel::geom_text_repel(
                aes(label = .data$sample_id),
                min.segment.length = 0,
                position = jit_pos
            )
        } else {
            p <- p + geom_text(
                aes(label = .data$sample_id),
                hjust = -0.1, vjust = 0.1,
                position = jit_pos
            )
        }
    }
    
    ## Y-axis
    y_label <- if (assay == "counts" & normalized) {
        "Normalized counts"
    } else if (assay == "counts") {
        "Counts"
    } else if (assay == "abundance") {
        "TPM - Transcripts Per Million"
    } else assay
    
    if (transform) {
        p <- p + scale_y_log10(name = paste0(y_label, " (log10 scale)"))
    } else {
        p <- p + scale_y_continuous(name = y_label)
    }
    
    p <- p + labs(title = paste0(genesymbol, " - ", gene))
    
    return(p)
}

```

## Checking the gene pattern across doses

```{r}
anno_df <- get_annotation_orgdb(
  de_container = dds_LR,
  orgdb_package = "org.Hs.eg.db",
  id_type = "ENSEMBL"
)


gene_plot <- plot_gene_modified(
  dds = dds_LR,
  gene = c("ENSG00000008517"),
  intgroup = "Dose",
  annotation_obj = anno_df
) + ggpt

gene_plot

# plot_gene_modified(
#      dds = dds_LR,
#      gene = c("ENSG00000008517"),
#      intgroup = "Dose",
#      annotation_obj = anno_df, dose_colors = c("#2166AC", "#B2182B")
#  ) 

# ggsave("cd83.png", plot = gene_plot, width = 8, height = 5, dpi = 400)
```

## plotting transcripts dose-response

```{r}

LR_Tx_counts_full <- read.table(file = "./Isoquant/OUT/OUT.transcript_model_grouped_counts.tsv")

tx_mat <- LR_Tx_counts_full
rownames(tx_mat) <- tx_mat$V1
tx_mat <- tx_mat[, -1]

colnames(tx_mat) <- c("0Gy-do1", "0Gy-do2", "0Gy-do3", "4Gy-do1", "4Gy-do2", "4Gy-do3")


dds_tx <- DESeqDataSetFromMatrix(
  countData = tx_mat,
  colData = LR_metadata,
  design = ~ Dose + Donors
)

dds_tx<- BIUMmisc::dds_gencode_to_ensembl(dds_tx)
# ENST00000440815 ENST00000325568

p1 <- plot_gene_modified(
  dds = dds_tx,
  gene = c("ENST00000325568"),
  intgroup = "Dose",
  annotation_obj = anno_df
) +
  labs(
    title = "IL-32 - ENST00000325568",
    y= "Normalized counts (log10 scale)",
    x = "Dose"
  ) + ggpt+ scale_y_continuous(
    limits = c(0, 30),
    expand = expansion(mult = 0)
  ) + theme(legend.position = "none")


p2 <- plot_gene_modified(
  dds = dds_tx,
  gene = c("ENST00000440815"),
  intgroup = "Dose",
  annotation_obj = anno_df
) +
  labs(
    title = "IL-32 - ENST00000440815",
    y= "Normalized counts (log10 scale)",
    x = "Dose"
  ) + ggpt+ scale_y_continuous(
    limits = c(0, 200),
    expand = expansion(mult = 0)
  ) + theme(legend.position = "none")


 p1 +p2
 
 
# 
# # having them in one plot with different data point shapes for each transcript
# 
# # Extract data for both transcripts
# df1 <- plot_gene_modified(
#   dds = dds_tx,
#   gene = "ENST00000325568",
#   intgroup = "Dose",
#   annotation_obj = anno_df,
#   return_data = TRUE  # <-- we'll need the raw data; you may need to modify your function to support this
# ) %>% mutate(transcript = "ENST00000325568")
# 
# df2 <- plot_gene_modified(
#   dds = dds_tx,
#   gene = "ENST00000440815",
#   intgroup = "Dose",
#   annotation_obj = anno_df,
#   return_data = TRUE
# ) %>% mutate(transcript = "ENST00000440815")
# 
# # Combine
# combined_df <- bind_rows(df1, df2)
# 
# 
# ggplot(combined_df, aes(x = Dose, y = exp_value, shape = transcript, color = Donors)) +
#   geom_point(size = 4) +  # bigger dots
#   scale_y_continuous(limits = c(0, 200), expand = expansion(mult = 0)) +
#   labs(
#     y = "Normalized counts (log10 scale)",
#     x = "Dose",
#     shape = "Transcript"
#   ) +
#   theme_classic() + ggpt


```

# Differential Expression Analysis- EdgeR

## gene level DEA

```{r}
gene_mat <- counts(dds_LR)

LR_samples <- data.frame(
  sample_id = LR_metadata$sample_id,
  group = LR_metadata$Dose,
  Donors = LR_metadata$Donors
)

knitr::kable(LR_samples)

# Create DGEList
y_DEA <- DGEList(
  counts = gene_mat,
  samples = LR_samples
)

keep <- filterByExpr(y_DEA)
y_DEA <- y_DEA[keep, keep.lib.sizes = FALSE]

# TMM normalization
y_DEA <- normLibSizes(y_DEA)

# Define factors
Donor <- factor(LR_metadata$Donors)
Treatment <- factor(LR_metadata$Dose)

# Design matrix including donor
design <- model.matrix(~ Donor + Treatment)

#  Estimate dispersions

y_DEA <- estimateDisp(y_DEA, design)

y_DEA$common.dispersion

# dispersion estimate plot

plotBCV(y_DEA)

#  Fit GLM model
fit_gene <- glmQLFit(y_DEA, design)

# plot quasi-likelihood despersion

plotQLDisp(fit_gene)

# Test for a treatment effect
lrt_gene <- glmQLFTest(fit_gene, coef = "Treatment4Gy")

# Extract results
edger_gene_res <- topTags(lrt_gene, n = Inf)


sig_genes_edger <- edger_gene_res$table[edger_gene_res$table$FDR <= 0.05, ]

# plot raw pvalues histogram

hist(edger_gene_res$table$PValue)


sig_genes_edger$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = rownames(sig_genes_edger),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```

## Volcanoplot

```{r}

de_res <- edger_gene_res$table

top_genes <- de_res[order(de_res$FDR), ][1:10, ]
top_genes$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = rownames(top_genes),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)


# Define up/down/NS
de_res$diffexpressed <- factor(ifelse(de_res$FDR < 0.05 & de_res$logFC > 0, "Up",
                                      ifelse(de_res$FDR < 0.05 & de_res$logFC < 0, "Down", "NS")
))

# Volcano plot
g <- ggplot(de_res, aes(x = logFC, y = -log10(FDR), color = diffexpressed)) +
  geom_point(alpha = 0.4, size = 3) +
  geom_text_repel(
    data = top_genes,
    aes(x = logFC, y = -log10(FDR), label = SYMBOL),
    size = 4,
    color = "black"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  xlim(c(-7, 7)) +
  xlab("log2 fold change") +
  ylab("-log10 FDR") +
  scale_color_manual(values = c("blue", "grey", "red")) +
  theme_bw() +
  theme(legend.position = "none") +
  ggpt

g

```

## transcript level DEA

```{r}
tx_mat <- LR_Tx_counts

y_DEA_tx <- DGEList(
  counts = tx_mat,
  samples = LR_samples
)

keep <- filterByExpr(y_DEA_tx)
y_DEA_tx <- y_DEA_tx[keep, , keep.lib.sizes = FALSE]

# TMM normalization
y_DEA_tx <- normLibSizes(y_DEA_tx)

# Define factors
Donor <- factor(LR_metadata$Donors)
Treatment <- factor(LR_metadata$Dose)

# Design matrix including donor
design <- model.matrix(~ Donor + Treatment)

#  Estimate dispersions
y_DEA_tx <- estimateDisp(y_DEA_tx, design)

y_DEA_tx$common.dispersion

# dispersion estimate plot

plotBCV(y_DEA_tx)

# Fit GLM model
fit_tx <- glmQLFit(y_DEA_tx, design)

# plot quasi-likelihood despersion

plotQLDisp(fit_tx)

# Test for a treatment effect

lrt_tx <- glmQLFTest(fit_tx, coef = "Treatment4Gy")

# Extract results
edger_Tx_res <- topTags(lrt_tx, n = Inf)


sig_tx <- edger_Tx_res$table[edger_Tx_res$table$FDR <= 0.05, ]

# Filter rows starting with "transcript"
sig_novel_tx <- sig_tx[grepl("^transcript", rownames(sig_tx)), ]

# plot raw pvalues histogram

hist(edger_Tx_res$table$PValue)
```

# diffsplice- DTU

## Preparing the matrix for the analysis

```{r}
# --- Set proper rownames once, from the ID column ---
LR_Tx_counts <- read.table(file = "./Isoquant/OUT/OUT.transcript_model_grouped_counts.tsv")
rownames(LR_Tx_counts) <- LR_Tx_counts$V1
LR_Tx_counts$V1 <- NULL

# --- Expression filter applies to ALL transcripts (annotated + novel) ---
LR_Tx_counts <- LR_Tx_counts[rowSums(LR_Tx_counts) > 6, ]

# --- Clean IDs: strip version suffix ONLY from Ensembl-style IDs; leave novel IDs untouched ---
LR_Tx_counts_filtered <- LR_Tx_counts
clean_ids <- rownames(LR_Tx_counts_filtered)
is_ensembl <- grepl("^ENST", clean_ids)
clean_ids[is_ensembl] <- sub("\\..*$", "", clean_ids[is_ensembl])
rownames(LR_Tx_counts_filtered) <- clean_ids

# --- Naming the columns ---
colnames(LR_Tx_counts_filtered) <- c("0Gy-do1", "0Gy-do2", "0Gy-do3",
                                      "4Gy-do1", "4Gy-do2", "4Gy-do3")


LR_Tx_counts_filtered <- as.data.frame(LR_Tx_counts_filtered)
LR_Tx_counts_filtered$feature_id <- rownames(LR_Tx_counts_filtered)


# --- 1. BioMart mapping for known transcripts (do this FIRST) ---
mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

transcript_to_gene <- getBM(
  attributes = c("ensembl_transcript_id",
                 "ensembl_gene_id",
                 "external_gene_name",
                 "description"),
  filters = "ensembl_transcript_id",
  values  = LR_Tx_counts_filtered$feature_id,
  mart    = mart
)

# --- 2. Mapping for novel transcripts from tmap ---
tmap_26$ref_gene_id_clean <- sub("\\..*$", "", tmap_26$ref_gene_id)
tmap_26$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys      = tmap_26$ref_gene_id_clean,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

novel_tx2gene <- tmap_26 %>%
  dplyr::filter(!is.na(ref_gene_id)) %>%   # drop the one novel-locus transcript with no ref gene (transcript61117)
  dplyr::transmute(
    ensembl_transcript_id = qry_id,
    ensembl_gene_id        = ref_gene_id_clean,   # <-- key change: cleaned Ensembl gene ID, not SYMBOL-dependent
    external_gene_name     = SYMBOL,               # can stay NA — cosmetic only, never filtered on
    description             = NA_character_
  )

# --- 3. Combine: BioMart takes priority for known tx, novel fills the rest ---
transcript_to_gene_full <- dplyr::bind_rows(transcript_to_gene, novel_tx2gene) %>%
  dplyr::distinct(ensembl_transcript_id, .keep_all = TRUE)

# --- 4. Single merge into the count matrix ---

LR_Tx_counts_filtered <- merge(
  LR_Tx_counts_filtered, transcript_to_gene_full,
  by.x = "feature_id", by.y = "ensembl_transcript_id",
  all.x = TRUE
)

# --- 5. Rename/reorder by name, not fragile position ---
LR_Tx_counts_filtered <- dplyr::rename(LR_Tx_counts_filtered, gene_id = ensembl_gene_id)

sample_cols <- setdiff(
  colnames(LR_Tx_counts_filtered),
  c("feature_id", "gene_id", "external_gene_name", "description")
)
LR_Tx_counts_filtered <- LR_Tx_counts_filtered[, c("feature_id", "gene_id", sample_cols)]

# --- 6. Only drop transcripts diffSplice truly can't use: missing gene_id ---
LR_Tx_counts_filtered <- LR_Tx_counts_filtered[!is.na(LR_Tx_counts_filtered$gene_id), ]

```

## running diffsplice

```{r}
mat <- as.matrix(LR_Tx_counts_filtered[, -c(1, 2)])
rownames(mat) <- LR_Tx_counts_filtered$feature_id

# 
# transcript_to_gene$feature_id <- transcript_to_gene$ensembl_transcript_id 
# transcript_to_gene$gene_id <- transcript_to_gene$ensembl_gene_id


gene_df <- LR_Tx_counts_filtered[, c(1, 2)]


y <- DGEList(counts = mat, samples = LR_samples, genes = gene_df)

keep <- filterByExpr(y, group = LR_metadata$Dose)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- normLibSizes(y)
y$samples

design <- model.matrix(~ Donors + group, data = y$samples)

fit <- glmQLFit(y, design)

ex <- diffSplice(
  fit = fit,
  coef = 4,
  geneid = "gene_id",
  exonid = "feature_id",
  robust = TRUE
)

ts <- topSplice(ex, number = Inf)
ts.tx <- topSplice(ex, test = "t", number = Inf)
```

### StageR

```{r}
pScreen <- ts$P.Value
names(pScreen) <- ts$gene_id
pConfirmation <- matrix(ts.tx$P.Value, ncol = 1)
rownames(pConfirmation) <- ts.tx$feature_id

tx2gene <- ts.tx[, c(1, 2)]

stageRObj2 <- stageRTx(
  pScreen = pScreen,
  pConfirmation = pConfirmation,
  pScreenAdjusted = FALSE,
  tx2gene = tx2gene
)

stageRObj2 <- stageWiseAdjustment(
  object = stageRObj2,
  method = "dtu",
  alpha = 0.05
)
suppressWarnings({
  ds.padj <- getAdjustedPValues(stageRObj2,
                                order = TRUE,
                                onlySignificantGenes = TRUE
  )
})
head(ds.padj)

ds.filtered <- ds.padj[ds.padj$transcript <= 0.05, ]

# Keep only transcript_id and external_gene_name
mapping <- transcript_to_gene %>%
  dplyr::select(ensembl_transcript_id, external_gene_name)

colnames(ds.filtered)[2] <- "ensembl_transcript_id"

# Join with your filtered dataframe
ds.filtered <- ds.filtered %>%
  left_join(mapping, by = "ensembl_transcript_id") %>%
  rename(gene_name = external_gene_name)



plotSplice(ex, geneid = "ENSG00000133316", genecolname = "gene_id", FDR = 0.05)

```

```{r}
library(edgeR)
library(dplyr)
library(stageR)
library(biomaRt)

## ============================================================
## 1. Load raw counts and do basic cleanup
## ============================================================


LR_Tx_counts_filtered <- as.data.frame(LR_Tx_counts)
LR_Tx_counts_filtered$feature_id <- rownames(LR_Tx_counts_filtered)

is_novel <- grepl("^transcript", LR_Tx_counts_filtered$feature_id)

# Strip GENCODE version suffix ONLY for annotated (ENST...) transcripts
LR_Tx_counts_filtered$feature_id[!is_novel] <-
  gsub("\\..*", "", LR_Tx_counts_filtered$feature_id[!is_novel])

cat("Annotated transcripts:", sum(!is_novel), "\n")
cat("Novel transcripts:", sum(is_novel), "\n")

## ============================================================
## 2. Gene mapping: BioMart for ALL known transcripts + tmap_novel_coding
##    for ONLY the 8 candidate novel transcripts
## ============================================================

mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

annotated_ids <- unique(LR_Tx_counts_filtered$feature_id[!is_novel])

transcript_to_gene <- getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_transcript_id",
  values = annotated_ids,
  mart = mart
) %>%
  dplyr::rename(
    feature_id      = ensembl_transcript_id,
    gene_id         = ensembl_gene_id,
    external_gene_name = external_gene_name
  )

novel_tx2gene <- tmap_novel_coding %>%
  dplyr::filter(!is.na(ref_gene_id)) %>%
  dplyr::transmute(
    feature_id         = qry_id,
    gene_id            = gsub("\\..*", "", ref_gene_id),
    external_gene_name = SYMBOL
  )

transcript_to_gene_full <- dplyr::bind_rows(transcript_to_gene, novel_tx2gene) %>%
  dplyr::distinct(feature_id, .keep_all = TRUE)

cat("Total gene mapping rows:", nrow(transcript_to_gene_full), "\n")
cat("Novel transcripts with a gene mapping:",
    sum(grepl("^transcript", transcript_to_gene_full$feature_id)), "\n")

## ============================================================
## 3. Merge counts with gene mapping
## ============================================================

LR_Tx_counts_filtered <- merge(LR_Tx_counts_filtered, transcript_to_gene_full,
                               by = "feature_id", all.x = TRUE)

LR_Tx_counts_filtered <- LR_Tx_counts_filtered[, c("feature_id", "gene_id.x",
                                                    "0Gy-do1", "0Gy-do2", "0Gy-do3",
                                                    "4Gy-do1", "4Gy-do2", "4Gy-do3")]
colnames(LR_Tx_counts_filtered)<- c("feature_id", "gene_id",
                                                    "0Gy-do1", "0Gy-do2", "0Gy-do3",
                                                    "4Gy-do1", "4Gy-do2", "4Gy-do3")
n_before <- nrow(LR_Tx_counts_filtered)
LR_Tx_counts_filtered <- na.omit(LR_Tx_counts_filtered)
n_after <- nrow(LR_Tx_counts_filtered)
cat("Rows dropped due to no gene mapping:", n_before - n_after, "\n")

stopifnot("transcript13826.chr15.nnic" %in% LR_Tx_counts_filtered$feature_id)
print(LR_Tx_counts_filtered[LR_Tx_counts_filtered$feature_id == "transcript13826.chr15.nnic", ])

## ============================================================
## 4. DGEList + diffSplice
## ============================================================

mat <- as.matrix(LR_Tx_counts_filtered[, -c(1, 2)])
rownames(mat) <- LR_Tx_counts_filtered$feature_id
gene_df <- LR_Tx_counts_filtered[, c("feature_id", "gene_id")]

y <- DGEList(counts = mat, samples = LR_samples, genes = gene_df)
keep <- filterByExpr(y, group = LR_metadata$Dose)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- normLibSizes(y)

cat("Transcripts surviving filterByExpr:", sum(keep), "\n")
cat("Novel transcripts surviving filterByExpr:",
    sum(grepl("^transcript", rownames(y))), "\n")
cat("transcript13826.chr15.nnic passed filterByExpr:",
    "transcript13826.chr15.nnic" %in% rownames(y), "\n")

design <- model.matrix(~ Donors + group, data = y$samples)
fit <- glmQLFit(y, design)

ex <- diffSplice(fit = fit, coef = 4, geneid = "gene_id", exonid = "feature_id", robust = TRUE)
ts    <- topSplice(ex, number = Inf)
ts.tx <- topSplice(ex, test = "t", number = Inf)

## ============================================================
## 5. stageR two-stage testing
## ============================================================

pScreen <- ts$P.Value
names(pScreen) <- ts$gene_id

pConfirmation <- matrix(ts.tx$P.Value, ncol = 1)
rownames(pConfirmation) <- ts.tx$feature_id

tx2gene_stageR <- ts.tx[, c("feature_id", "gene_id")]

stageRObj2 <- stageRTx(pScreen = pScreen, pConfirmation = pConfirmation,
                      pScreenAdjusted = FALSE, tx2gene = tx2gene_stageR)
stageRObj2 <- stageWiseAdjustment(object = stageRObj2, method = "dtu", alpha = 0.05)

suppressWarnings({
  ds.padj <- getAdjustedPValues(stageRObj2, order = TRUE, onlySignificantGenes = TRUE)
})

## Inspect actual column names before renaming anything further
cat("\nds.padj column names:\n")
print(colnames(ds.padj))

## ============================================================
## 6. Add gene symbols for readability
## ============================================================

# ds.padj typically has columns: geneID, txID, gene (p-value), transcript (p-value)
# Rename explicitly by matching known stageR output names, not by position
ds.padj <- ds.padj %>%
  dplyr::rename(
    gene_id_stageR = geneID,
    feature_id     = txID
  )

gene_symbol_lookup <- transcript_to_gene_full %>%
  dplyr::distinct(gene_id, external_gene_name)

ds.filtered <- ds.padj %>%
  dplyr::left_join(gene_symbol_lookup, by = c("gene_id_stageR" = "gene_id")) %>%
  dplyr::rename(gene_name = external_gene_name)

ds.significant <- ds.filtered[ds.filtered$transcript <= 0.05, ]

## ============================================================
## 7. Pull the FBXO22 candidate result specifically
## ============================================================

fbxo22_result <- ds.filtered[ds.filtered$feature_id == "transcript13826.chr15.nnic", ]
print(fbxo22_result)

raw_result <- ts.tx[ts.tx$feature_id == "transcript13826.chr15.nnic", ]
print(raw_result)
```

### creating the function for DTU plotting

```{r}
plot_dtu_transcripts <- function(
  gene_id,
  LR_Tx_counts = LR_Tx_counts_filtered,
  dds = dds_tx,
  annotation_df = NULL,
  exclude_transcripts = NULL,
  transform = TRUE,
  dtu_result, 
  dose_samples = list(
    "0Gy" = c("0Gy-do1", "0Gy-do2", "0Gy-do3"),
    "4Gy" = c("4Gy-do1", "4Gy-do2", "4Gy-do3")
  ),
  log_transform = FALSE,
  title = NULL
) {

  
  gene_id_arg <- gene_id
  
  gene_label <- gene_id_arg

if (!is.null(annotation_df)) {
  if (!all(c("gene_id", "gene_name") %in% colnames(annotation_df))) {
    stop("`annotation_df` must contain columns: gene_id, gene_name")
  }

  mapped_name <- annotation_df %>%
    dplyr::filter(gene_id == gene_id_arg) %>%
    dplyr::pull(gene_name) %>%
    unique()

  if (length(mapped_name) >= 1 && !is.na(mapped_name[1])) {
    gene_label <- mapped_name[1]
  }
}


  ## ---- 1) transcripts for gene ----
  tx_ids <- LR_Tx_counts %>%
    dplyr::filter(gene_id == gene_id_arg) %>%
    dplyr::pull(feature_id)
  
  if (!is.null(exclude_transcripts)) {
  tx_ids <- setdiff(tx_ids, exclude_transcripts)

  if (length(tx_ids) == 0) {
    stop("All transcripts were excluded for gene: ", gene_id_arg)
  }
}

  if (length(tx_ids) == 0) {
    stop("No transcripts found for gene: ", gene_id_arg)
  }

  ## ---- 2) normalized counts ----
  norm_counts <- counts(dds, normalized = TRUE)[tx_ids, , drop = FALSE]

  if (isTRUE(log_transform)) {
    norm_counts <- log10(norm_counts + 1)
  }

  ## ---- 3) long dataframe ----
  df <- as.data.frame(norm_counts) %>%
    tibble::rownames_to_column("transcript") %>%
    tidyr::pivot_longer(
      cols = -transcript,
      names_to = "sample",
      values_to = "normalized_counts"
    )

  ## ---- 4) dose info ----
  df <- df %>%
    dplyr::mutate(
      Dose = dplyr::case_when(
        sample %in% dose_samples[["0Gy"]] ~ "0Gy",
        sample %in% dose_samples[["4Gy"]] ~ "4Gy",
        TRUE ~ NA_character_
      )
    )

  ## ---- 5) DTU significance ----
  sig_tx <- unique(dtu_result$ensembl_transcript_id)

  df <- df %>%
    dplyr::mutate(
      DTU_status = ifelse(
        transcript %in% sig_tx,
        "Significant",
        "Not_significant"
      ),
      Fill_group = interaction(Dose, DTU_status, drop = TRUE)
    )

  ## ---- 6) transcript order ----
  transcript_order <- df %>%
    dplyr::group_by(transcript) %>%
    dplyr::summarise(total = sum(normalized_counts, na.rm = TRUE)) %>%
    dplyr::arrange(dplyr::desc(total)) %>%
    dplyr::pull(transcript)

  df$transcript <- factor(df$transcript, levels = transcript_order)

  ## ---- 7) title ----
  if (is.null(title)) {
    title <- paste0("DTU transcripts for gene ", gene_label)
    if (isTRUE(log_transform)) title <- paste0(title, " (log10+1)")
  }


  ## ---- 8) colors ----
  fill_colors <- c(
    "0Gy.Significant"     = "#2166AC",  # blue
    "4Gy.Significant"     = "#B2182B",  # red
    "0Gy.Not_significant" = "grey80",
    "4Gy.Not_significant" = "grey40"
  )

  ## ---- 9) plot ----
  p <- ggplot2::ggplot(
    df,
    ggplot2::aes(
      x = transcript,
      y = normalized_counts,
      fill = Fill_group,
      group = sample
    )
  ) +
    ggplot2::geom_bar(
      stat = "identity",
      position = ggplot2::position_dodge(width = 0.9)
    ) +
    ggplot2::scale_fill_manual(
      values = fill_colors,
      breaks = c(
        "0Gy.Significant",
        "4Gy.Significant",
        "0Gy.Not_significant",
        "4Gy.Not_significant"
      ),
      labels = c(
        "0Gy (DTU)",
        "4Gy (DTU)",
        "0Gy (not DTU)",
        "4Gy (not DTU)"
      ),
      name = "Dose / DTU status"
    ) +
    ggplot2::labs(
      x = "Transcript ID",
      y = ifelse(log_transform,
                 "Normalized counts (log10+1)",
                 "Normalized counts"),
      title = title
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
  
  p <- p + theme(
    plot.title   = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold"),
    axis.text.x  = element_text(angle = 45, hjust = 1)
  )
  
  # y_label <- "normalized_counts"
  #   if (transform) {
  #   p <- p + scale_y_log10(name = paste0(y_label, "(log10 scale)"))
  # } else {
  #   p <- p + scale_y_continuous(name = y_label)
  # }
  # 


  return(p)
}

```

```{r}
plot_dtu_transcripts <- function(
        gene_id,
        LR_Tx_counts= LR_Tx_counts_filtered,
        dds =dds_tx,
        dtu_result,
        annotation_df = NULL,
        exclude_transcripts = NULL,
        dose_samples = list(
            "0Gy" = c("0Gy-do1", "0Gy-do2", "0Gy-do3"),
            "4Gy" = c("4Gy-do1", "4Gy-do2", "4Gy-do3")
        ),
        log_transform = FALSE,
        title = NULL,
        transform = TRUE
) {
    
    ## ---- packages ----
    required_pkgs <- c("dplyr", "tibble", "tidyr", "ggplot2", "ggpattern")
    missing_pkgs <- required_pkgs[
        !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
    ]
    if (length(missing_pkgs) > 0) {
        stop("Please install required packages: ",
             paste(missing_pkgs, collapse = ", "))
    }
    
    gene_id_arg <- gene_id
    
    ## ---- gene label (optional mapping) ----
    gene_label <- gene_id_arg
    if (!is.null(annotation_df)) {
        if (!all(c("gene_id", "gene_name") %in% colnames(annotation_df))) {
            stop("`annotation_df` must contain columns: gene_id, gene_name")
        }
        
        mapped <- annotation_df %>%
            dplyr::filter(gene_id == gene_id_arg) %>%
            dplyr::pull(gene_name) %>%
            unique()
        
        if (length(mapped) > 0 && !is.na(mapped[1])) {
            gene_label <- mapped[1]
        }
    }
    
    ## ---- 1) transcripts for gene ----
    tx_ids <- LR_Tx_counts %>%
        dplyr::filter(gene_id == gene_id_arg) %>%
        dplyr::pull(feature_id)
    
    if (!is.null(exclude_transcripts)) {
        tx_ids <- setdiff(tx_ids, exclude_transcripts)
    }
    
    if (length(tx_ids) == 0) {
        stop("No transcripts left for gene: ", gene_id_arg)
    }
    
    ## ---- 2) normalized counts ----
    norm_counts <- counts(dds, normalized = TRUE)[tx_ids, , drop = FALSE]
    
    if (log_transform) {
        norm_counts <- log10(norm_counts + 1)
    }
    
    ## ---- 3) long dataframe ----
    df <- as.data.frame(norm_counts) %>%
        tibble::rownames_to_column("transcript") %>%
        tidyr::pivot_longer(
            cols = -transcript,
            names_to = "sample",
            values_to = "value"
        )
    
    ## ---- 4) dose info ----
    df <- df %>%
        dplyr::mutate(
            Dose = dplyr::case_when(
                sample %in% dose_samples[["0Gy"]] ~ "0Gy",
                sample %in% dose_samples[["4Gy"]] ~ "4Gy",
                TRUE ~ NA_character_
            )
        ) %>%
        dplyr::filter(!is.na(Dose))
    
    ## ---- 5) DTU status ----
    sig_tx <- unique(dtu_result$ensembl_transcript_id)
    
    df <- df %>%
        dplyr::mutate(
            DTU_status = ifelse(
                transcript %in% sig_tx,
                "DTU",
                "non-DTU"
            ),
            Fill_group = interaction(Dose, DTU_status, drop = TRUE)
        )
    
    ## ---- 6) transcript order ----
    transcript_order <- df %>%
        dplyr::group_by(transcript) %>%
        dplyr::summarise(total = sum(value), .groups = "drop") %>%
        dplyr::arrange(dplyr::desc(total)) %>%
        dplyr::pull(transcript)
    
    df$transcript <- factor(df$transcript, levels = transcript_order)
    
    ## ---- 7) title ----
    if (is.null(title)) {
        title <- paste0("DTU transcripts for ", gene_label)
        if (log_transform) title <- paste0(title, " (log10+1)")
    }
    
    ## ---- 8) colors ----
    fill_colors <- c(
        "0Gy.DTU"     = "#2166AC",
        "4Gy.DTU"     = "#B2182B",
        "0Gy.non-DTU" = "grey80",
        "4Gy.non-DTU" = "grey40"
    )
    
    ## ---- 9) split data (PERFORMANCE SAFE) ----
    df_dtu    <- df %>% dplyr::filter(DTU_status == "DTU")
    df_nondtu <- df %>% dplyr::filter(DTU_status == "non-DTU")
    
    ## ---- 10) plot ----
    p <- ggplot2::ggplot() +
        
        ## non-DTU (solid grey)
        ggplot2::geom_bar(
            data = df_nondtu,
            ggplot2::aes(
                x = transcript,
                y = value,
                fill = Fill_group,
                group = sample
            ),
            stat = "identity",
            position = ggplot2::position_dodge(width = 0.9),
            colour = "black"
        ) +
        
        ## DTU (colored + striated)
        ggpattern::geom_bar_pattern(
            data = df_dtu,
            ggplot2::aes(
                x = transcript,
                y = value,
                fill = Fill_group,
                group = sample
            ),
            stat = "identity",
            position = ggplot2::position_dodge(width = 0.9),
            colour = "black",
            pattern = "stripe",
            pattern_fill = "black",
            pattern_angle = 45,
            pattern_density = 0.3,
            pattern_spacing = 0.05
        ) +
        
        ggplot2::scale_fill_manual(
            values = fill_colors,
            breaks = c(
                "0Gy.DTU",
                "4Gy.DTU",
                "0Gy.non-DTU",
                "4Gy.non-DTU"
            ),
            labels = c(
                "0Gy (DTU)",
                "4Gy(DTU)",
                "0Gy(non-DTU)",
                "4Gy(non-DTU)"
            ),
            name = "Transcript status"
        ) +
        
        ggplot2::labs(
            x = "Transcript ID",
            y = ifelse(log_transform,
                       "Normalized counts (log10+1)",
                       "Normalized counts"),
            title = title
        ) +
        
        ggplot2::theme_classic() +
        ggplot2::theme(
            plot.title   = ggplot2::element_text(size = 16, face = "bold", hjust = 0.5),
            axis.title.x = ggplot2::element_text(size = 13, face = "bold"),
            axis.title.y = ggplot2::element_text(size = 13, face = "bold"),
            axis.text.x  = ggplot2::element_text(angle = 45, hjust = 1),
            legend.position = "top"
        )
    
      y_label <- "normalized_counts"
    if (transform) {
    p <- p + scale_y_log10(name = paste0(y_label, "(log10 scale)"))
  } else {
    p <- p + scale_y_continuous(name = y_label)
  }
      
    return(p)
  

}

```

### plot DTU results
```{r}

dds_tx<- DESeq(dds_tx)

plot_dtu_transcripts("transcript13826.chr15.nnic", dtu_result = ds.filtered, annotation_df = anno_df)

plot_dtu_transcripts(gene_id = "ENSG00000105372", exclude_transcripts = c("ENST00000221975", "ENST00000598742") , dtu_result = ds.filtered ,annotation_df = anno_df )

```


# Enrichment --> Kegg, GO, reactome

```{r}
deg_genes <- rownames(sig_genes_edger)

# Convert gene symbols to Entrez IDs
entrez_genes <- bitr(deg_genes,
                     fromType = "ENSEMBL",
                     toType = "ENTREZID",
                     OrgDb = org.Hs.eg.db
)

# Perform KEGG pathway enrichment
kegg_res <- enrichKEGG(
  gene = entrez_genes$ENTREZID,
  organism = "hsa", # "hsa" for human
  pvalueCutoff = 0.05
)

kegg_res <- kegg_res@result[kegg_res@result$p.adjust <= 0.05, ]


# reactome pathway analysis

reactome_res <- enrichPathway(
  gene = entrez_genes$ENTREZID,
  organism = "human",
  pvalueCutoff = 0.05,
  readable = TRUE
)

reactome_res <- reactome_res@result[reactome_res@result$p.adjust <= 0.05, ]

# GO enrichment

deg_genes_symbol <- mapIds(
  org.Hs.eg.db,
  keys = deg_genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

clusterprofiler_res <- enrichGO(deg_genes_symbol,
                                keyType = "SYMBOL",
                                ont = "BP",
                                OrgDb = "org.Hs.eg.db",
                                pvalueCutoff = 0.05,
                                pAdjustMethod = "BH",
                                qvalueCutoff = 0.2,
                                minGSSize = 10,
                                maxGSSize = 500
)


clusterprofiler_res_sig <- clusterprofiler_res@result[clusterprofiler_res@result$p.adjust <= 0.05, ]

View(clusterprofiler_res_sig)
```

```{r}

deg_genes <- rownames(sig_genes_edger)

# Convert gene symbols to Entrez IDs
entrez_genes <- bitr(deg_genes,
                     fromType = "ENSEMBL",
                     toType = "ENTREZID",
                     OrgDb = org.Hs.eg.db
)

# the exact tested universe — same object used throughout our edgeR/DESeq2 comparisons
tested_genes_ensembl <- rownames(y_DEA)   # 13,020 genes


# map to SYMBOL and ENTREZID for the three different enrichment functions
universe_symbol <- mapIds(org.Hs.eg.db, keys = tested_genes_ensembl, column = "SYMBOL",
                           keytype = "ENSEMBL", multiVals = "first")
universe_symbol <- na.omit(universe_symbol)

universe_entrez <- bitr(tested_genes_ensembl, fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

universe_entrez_dedup <- universe_entrez[!duplicated(universe_entrez$ENSEMBL), ]
nrow(universe_entrez)        # check if inflated
nrow(universe_entrez_dedup)  # should be close to 13,020 (minus any truly unmappable genes)

entrez_genes_dedup <- entrez_genes[!duplicated(entrez_genes$ENSEMBL), ]  # one Entrez ID per input gene

universe_symbol_dedup <- universe_symbol[!duplicated(universe_symbol)]
length(universe_symbol_dedup)


# --- GO ---
clusterprofiler_res <- enrichGO(
  gene = deg_genes_symbol,
  universe = universe_symbol_dedup,
  keyType = "SYMBOL", ont = "BP", OrgDb = "org.Hs.eg.db",
  pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 0.2,
  minGSSize = 10, maxGSSize = 500
)
clusterprofiler_res_sig <- clusterprofiler_res@result[clusterprofiler_res@result$p.adjust <= 0.05, ]


# --- KEGG ---
kegg_res <- enrichKEGG(
  gene = entrez_genes_dedup$ENTREZID,      # deduplicated — 171 genes, not 1,043
  universe = universe_entrez_dedup$ENTREZID,      # correct filtered background
  organism = "hsa",
  pvalueCutoff = 0.05
)
kegg_res_sig <- kegg_res@result[kegg_res@result$p.adjust <= 0.05, ]
kegg_res_sig[, c("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust")]


# --- Reactome ---
reactome_res <- enrichPathway(
  gene = entrez_genes_dedup$ENTREZID,
  universe = universe_entrez_dedup$ENTREZID,
  organism = "human",
  pvalueCutoff = 0.05,
  readable = TRUE
)
reactome_res_sig <- reactome_res@result[reactome_res@result$p.adjust <= 0.05, ]

# specifically check whether the p53 pathway and the marginal immune term survive
kegg_res_sig[grepl("hsa04115", kegg_res_sig$ID), ]
clusterprofiler_res_sig[grepl("GO:0002478", clusterprofiler_res_sig$ID), ]


nrow(entrez_genes)                          # total rows returned by bitr()
length(unique(entrez_genes$ENSEMBL))        # how many distinct input genes actually mapped
sum(duplicated(entrez_genes$ENSEMBL))       # how many input genes mapped to >1 Entrez ID
nrow(entrez_genes_dedup)  # should now be ≤ 183
length(deg_genes)                           # should be 183 — your original input
                                    # 183, total DEGs
sum(!is.na(deg_genes_symbol))                         # how many got a SYMBOL
sum(is.na(deg_genes_symbol))                          # how many lack a gene symbol (R2 expects 12)
nrow(entrez_genes_dedup)                              # how many mapped to a unique Entrez ID (used for KEGG/Reactome)

```

## GeneTonic

```{r}
# create DESeq result object

de_res_table <- edger_gene_res$table

deseq_like <- DataFrame(
  log2FoldChange = de_res_table$logFC,
  stat = de_res_table$F,
  pvalue = de_res_table$PValue,
  padj = de_res_table$FDR,
  row.names = rownames(de_res_table)
)

deseq_like <- DESeqResults(deseq_like)

# create GeneTonic object

gtl_4Gy <- GeneTonicList(
  dds = dds_LR,
  res_de = deseq_like,
  res_enrich = shake_enrichResult(clusterprofiler_res),
  annotation_obj = anno_df
)

# GeneTonic(gtl= gtl_4Gy)


# preparing the matrix for heatmap

vsd_LR_corrected <- vsd_LR

assay(vsd_LR_corrected) <- limma::removeBatchEffect(assay(vsd_LR_corrected), batch = vsd_LR_corrected$Donors)
scored_mat <- GeneTonic::gs_scores(se = vsd_LR_corrected, gtl = gtl_4Gy)


anno <- as.data.frame(colData(vsd_LR_corrected)[, c("Sex", "Donors", "Dose")])

dose_colors <- colorRampPalette(c("#2166AC", "#B2182B"))(2)
names(dose_colors) <- c("0Gy", "4Gy")

anno_colors <- list(
  Sex = c(female = "pink", male = "deepskyblue3"),
  Dose = dose_colors
)


column_annotation <- HeatmapAnnotation(
  Sex = anno$Sex,
  Dose = anno$Dose,
  col = anno_colors, show_legend = T
)

library(circlize)

col_fun <- colorRamp2(
  c(-2, 0, 2),
  c("blue", "white", "red")
)

ht <- Heatmap(scored_mat[1:15, ],
              name = "Expression",
              top_annotation = column_annotation,
              cluster_columns = F, # Optional: Cluster columns
              cluster_rows = F, # Optional: Cluster rows
              show_row_names = T,
              show_column_names = TRUE,
              row_dend_side = "left",
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 10, fontface = "bold"),
              width = unit(7, "cm"),
              heatmap_height = unit(12, "cm"),
              column_names_gp = gpar(fontsize = 7),
              column_labels = rep(c(1, 2, 3), 2),
              column_names_rot = 360,
              column_title = "Donors",
              column_title_side = "bottom",
              show_heatmap_legend = T,
              col = col_fun
)


draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right")
```

## enrichment for the downregulated genes

```{r}
dwn_sig <- sig_genes_edger[sig_genes_edger$logFC < 0, ]

dwn_genes <- dwn_sig$SYMBOL

clusterprofiler_res_dwn <- enrichGO(dwn_genes,
                                    keyType = "SYMBOL",
                                    ont = "BP",
                                    OrgDb = "org.Hs.eg.db",
                                    pvalueCutoff = 0.05,
                                    pAdjustMethod = "BH",
                                    qvalueCutoff = 0.2,
                                    minGSSize = 10,
                                    maxGSSize = 500
)


clusterprofiler_res_dwn <- clusterprofiler_res_dwn@result[clusterprofiler_res_dwn@result$p.adjust <= 0.05, ]

View(clusterprofiler_res_dwn)
```

# Gene Set Variation Analysis-GSVA

```{r}
exprs_mat <- assay(dds_LR)

exprs_mat <- exprs_mat[!grepl("^transcript", rownames(exprs_mat)), ]


exprs_mat <- exprs_mat[rowSums(exprs_mat) > 6, ]


genes_exprs_mat <- rownames(exprs_mat)

genes_exprs_mat <- mapIds(
  org.Hs.eg.db,
  keys = genes_exprs_mat,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

exprs_df <- as.data.frame(exprs_mat)
exprs_df$SYMBOL <- genes_exprs_mat

# calculating the variance

exprs_df$variance <- apply(exprs_df[, !(names(exprs_df) %in% c("SYMBOL"))], 1, var, na.rm = TRUE)

# 2. Keep the most variable transcript per gene symbol
exprs_highvar <- exprs_df %>%
  group_by(SYMBOL) %>%
  slice_max(order_by = variance, n = 1, with_ties = FALSE) %>%
  ungroup()

# 3. Remove helper column
exprs_highvar <- exprs_highvar %>% dplyr::select(-variance)

# 4. Set SYMBOL as rownames
exprs_highvar <- as.data.frame(exprs_highvar)

exprs_highvar <- na.omit(exprs_highvar)
rownames(exprs_highvar) <- exprs_highvar$SYMBOL
exprs_highvar$SYMBOL <- NULL

exprs_highvar <- as.matrix(exprs_highvar)

# run GSVA

goannot <- AnnotationDbi::select(org.Hs.eg.db, keys = keys(org.Hs.eg.db, keytype = "SYMBOL"), columns = "GOALL", keytype = "SYMBOL")

genesbygo <- split(goannot$SYMBOL, goannot$GO)

genesbygo_dedup <- lapply(split(goannot$SYMBOL, goannot$GO), unique)


huangPar <- gsvaParam(exprs_highvar, genesbygo_dedup, minSize = 5, maxSize = 2000)

esmicro <- gsva(huangPar)

# adding a column for GO id description

go_ids <- rownames(esmicro)

# Get the term names
go_terms <- Term(GOTERM[go_ids])
# Add as a new column
esmicro <- as.data.frame(esmicro)
esmicro$Description <- go_terms
rownames(esmicro) <- esmicro$Description
esmicro <- esmicro[, -7]


# Applying Limma on GSVA matrix to extract differentially expressed pathways

# creating metadata

patient <- rep(c("p1", "p2", "p3"), 2)
dose <- c(rep("0Gy", 3), rep("4Gy", 3))
metadata <- data.frame(patient, dose)

# make the design

f <- factor(paste(metadata$dose))

design <- model.matrix(~ 0 + f)

patient <- as.factor(metadata$patient)

# limma-based DEA

corfit <- duplicateCorrelation(esmicro, design, block = patient)

fit <- lmFit(esmicro, design, block = patient, correlation = corfit$consensus)

contrast.matrix <- makeContrasts((f4Gy - f0Gy), levels = design)

fit2 <- contrasts.fit(fit, contrast.matrix)

fit2 <- eBayes(fit2)

results_gsva <- topTable(fit2, coef = 1, number = nrow(esmicro), adjust = "BH")

results_f_gsva <- decideTests(fit2, p.value = 0.05)

summary(results_f_gsva)


results_f_gsva <- as.data.frame(results_f_gsva)


sig_gsva <- results_f_gsva[results_f_gsva[, 1] == 1, ]
```

## GSEA using GO terms

```{r}
## background genes

background_genes <- edger_gene_res$table %>%
  as.data.frame() %>%
  filter(logCPM != 0) %>%
  tibble::rownames_to_column(var = "gene") %>%
  pull(gene)


res_df <- edger_gene_res$table %>%
  as.data.frame() %>%
  filter(logCPM != 0) %>%
  tibble::rownames_to_column(var = "gene")

background_genes_map <- bitr(
  geneID = background_genes,
  fromType = "ENSEMBL",
  toType = "SYMBOL",
  OrgDb = "org.Hs.eg.db"
)


m_df <- msigdbr(species = "Homo sapiens")

head(m_df)

go_df <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% dplyr::select(gs_name, gene_symbol)

### Gene set enrichment analysis

## rank all the genes by signed fold change * -log10pvalue.

res_df <- res_df %>%
  mutate(signed_rank_stats = sign(logFC) * -log10(PValue)) %>%
  left_join(background_genes_map, by = c("gene" = "ENSEMBL")) %>%
  arrange(desc(signed_rank_stats))

# removing duplicates by choosing the ones with FDR <=0.05

res_df_filtered <- res_df %>%
  dplyr::group_by(SYMBOL) %>%
  dplyr::arrange(FDR, .by_group = TRUE) %>%
  dplyr::slice( # keep logic per group
    if (any(FDR <= 0.05)) which.min(FDR) else 1
  ) %>%
  dplyr::ungroup()


res_df_filtered <- na.omit(res_df_filtered)

gene_list <- res_df_filtered$signed_rank_stats
names(gene_list) <- res_df_filtered$SYMBOL
gene_list <- sort(gene_list, decreasing = TRUE)


em2 <- GSEA(gene_list, TERM2GENE = go_df)

em2@result %>% View()

write.csv(em2@result, "GSEA_results.csv")

### visualization

p1 <- gseaplot(em2,
               geneSetID = "GOBP_PEPTIDE_ANTIGEN_ASSEMBLY_WITH_MHC_CLASS_II_PROTEIN_COMPLEX",
               by = "runningScore", title = "GOBP_PEPTIDE_ANTIGEN_ASSEMBLY_WITH_MHC_CLASS_II_PROTEIN_COMPLEX"
)

p2 <- gseaplot(em2,
               geneSetID = "GOBP_HUMORAL_IMMUNE_RESPONSE",
               by = "runningScore", title = "GOBP_HUMORAL_IMMUNE_RESPONSE"
)

p1 / p2

```

# Checking novel significant transcripts coding potential
## Transdecoder

```{r}

# Read protein FASTA file
pep <- readAAStringSet("novel_isoforms_n.fa.transdecoder.pep")

# Inspect
pep
names(pep)[1:5] # transcript IDs
width(pep)[1:5] # sequence lengths
pep[1] # first sequence

# convert to df
headers <- names(pep)
seqs <- as.character(pep)

# Parse each header
pep_df <- tibble(
  transcript_id = headers,
  sequence = seqs,
  orf_type = str_extract(headers, "(?<=ORF type:)\\w+"),
  strand = str_extract(headers, "(?<=\\()(\\+|\\-)\\)(?=,score=)"),
  score = str_extract(headers, "(?<=score=)[0-9.]+") %>% as.numeric(),
  length = str_extract(headers, "(?<=len:)\\d+") %>% as.numeric(),
  start = str_extract(headers, "(?<=:)\\d+(?=-)") %>% as.numeric(),
  end = str_extract(headers, "(?<=-)\\d+(?=\\([+-]\\))") %>% as.numeric(),
  coord_strand = str_extract(headers, "(?<=\\d\\()[+-](?=\\))")
)

pep_df <- pep_df %>%
  mutate(final_strand = ifelse(!is.na(coord_strand), coord_strand, strand)) %>%
  dplyr::select(-strand, -coord_strand)

# filter on the tx that are potentially coding genes

pep_df_strict <- pep_df %>%
  dplyr::filter(
    score >= 100,
    orf_type == "complete"
  )


pep_df$clean_id <- sub("^([^.]+\\.[^.]+\\.[^.]+).*", "\\1", pep_df[[1]])

pep_df_strict$clean_id <- sub("^([^.]+\\.[^.]+\\.[^.]+).*", "\\1", pep_df_strict[[1]])
```

## CPC2

```{r}
novel_isoforms_fa <- readAAStringSet("novel_isoforms_n.fa")



novel_isoforms_fa_df <- as.data.frame(novel_isoforms_fa)

# focusing on the 8 transcripts detected by transdecoder

selected_8_tx <- c("transcript13458.chr22.nnic", "transcript13760.chr9.nnic", "transcript13826.chr15", "transcript15213.chr10.nnic", "transcript21609.chr6.nnic", "transcript21664.chr9.nnic", "transcript23187.chr5.nnic", "transcript36523.chr12.nnic")

selected_tx_fa <- novel_isoforms_fa_df[selected_8_tx, ]


ids <- paste0("seq", seq_along(selected_tx_fa))

fa_file <- "selected_tx.fa"
writeLines(
  paste0(">", ids, "\n", selected_tx_fa),
  con = fa_file
)



cpc_res <- read.table("cpc_res.txt")
colnames(cpc_res) <- c("ID", "peptide_length", "Fickett_score", "pI", "ORF_integrity", "coding_probability", "label")

cpc_res_coding <- cpc_res[cpc_res$label == "coding", ]
cpc_res_noncoding <- cpc_res[cpc_res$label == "noncoding", ]

```

## CPAT

```{r}
cpat_res <- read_tsv("cpat_results.txt.ORF_prob.best.tsv")

cpat_res_coding <- cpat_res[cpat_res$Coding_prob >= 0.5, ]

View(cpat_res_coding)
```

### focusing only on the novel coding transcripts that are shared between all the methods

```{r}
venn.plot <- venn.diagram(
  x = list(
    TransDecoder = unique(pep_df$clean_id),
    CPC2 = cpc_res_coding$ID,
    CPAT = tolower(cpat_res_coding$seq_ID)
  ),
  filename = NULL, # Plot to R directly
  fill = c("#66c2a5", "#8da0cb", "orange"),
  alpha = 0.6,
  cat.cex = 1.2,
  cex = 1.5,
  #  main = "Novel coding isoforms"
)

grid::grid.draw(venn.plot)


TransDecoder <- tolower(unique(pep_df$clean_id))
CPC2 <- tolower(cpc_res_coding$ID)
CPAT <- tolower(cpat_res_coding$seq_ID)

shared_transcripts <- Reduce(intersect, list(TransDecoder, CPC2, CPAT))
length(shared_transcripts)

```

## Gffcompare

```{r}
tmap <- read.table("Isoquant/OUT/gffcmp_out.OUT.transcript_models.gtf.tmap",
                   sep = "\t", header = FALSE, quote = "", stringsAsFactors = FALSE, fill = TRUE
)

tmap_26

# 3. Filter the tmap file to keep only those novel transcripts


tmap_novel_coding <- tmap %>%
  filter(V5 %in% shared_transcripts)

colnames(tmap_novel_coding) <- tmap[1, ]


tmap_novel_coding[, 1] <- gsub("\\..*", "", tmap_novel_coding[, 1])

tmap_novel_coding$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = tmap_novel_coding$ref_gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```

### plotting the number of coding and non coding novel transcripts detected by each method

```{r}
# For CPAT: classify based on coding probability threshold
cpat_res_labeled <- cpat_res %>%
  mutate(
    tool = "CPAT",
    label = ifelse(Coding_prob >= 0.5, "Coding", "Non-coding")
  )

# For CPC2: assume the column is named 'coding' or similar
cpc_res_labeled <- cpc_res %>%
  mutate(
    tool = "CPC2",
    label = ifelse(tolower(label) == "coding", "Coding", "Non-coding")
  )

# --- TransDecoder ---
# there are 14 total transcripts (11 unique) with ORFs
# Let's get unique ones
transdecoder_unique <- unique(pep_df$clean_id)

# Total number of isoforms in your dataset (e.g., 26 total)
total_isoforms <- 26 # <-- change this to your real total number

# Compute counts
transdecoder_df <- tibble(
  tool = "TransDecoder",
  label = c("Coding", "Non-coding"),
  count = c(length(transdecoder_unique), total_isoforms - length(transdecoder_unique))
)


# Count per tool and label
summary_df <- bind_rows(cpat_res_labeled, cpc_res_labeled) %>%
  count(tool, label, name = "count")

# Add TransDecoder data
summary_df <- bind_rows(summary_df, transdecoder_df)

ggplot(summary_df, aes(x = tool, y = count, fill = label)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Coding" = "steelblue", "Non-coding" = "orange")) +
  labs(
    title = "Comparison of Coding vs Non-coding Transcripts Across Tools",
    x = "Tool",
    y = "Count",
    fill = "Transcript Type"
  ) +
  theme_minimal(base_size = 20) +  theme(
    plot.title   = element_text(size = 20, face = "bold"),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )

```

# Session info

```{r}
sessionInfo()
```


# revision

```{r}
library(tximport)

salmon_dir <- "salmon_quant_gcBias"

# restrict to the 0Gy and 4Gy, 6hrs samples only — matching the LR study design
target_samples <- c(
  "0Gy-6hrs-A_S23", "0Gy-6hrs-F_S43", "0Gy-6hrs-S_S33",
  "4Gy-6hrs-A_S27", "4Gy-6hrs-F_S47", "4Gy-6hrs-S_S37"
)

sample_names <- target_samples
files <- file.path(salmon_dir, sample_names, "quant.sf")
names(files) <- sample_names

# sanity check before importing
all(file.exists(files))

# SR metadata
sr_metadata <- data.frame(
  sample_id = sample_names,
  Dose = ifelse(grepl("^0Gy", sample_names), "0Gy", "4Gy"),
  Donors = sub(".*-(A|F|S)_S[0-9]+$", "\\1", sample_names)
)
sr_metadata

# import

# point to your quant.sf files, one per sample
files <- file.path("salmon_quant_gcBias", sample_names, "quant.sf")
names(files) <- sample_names

# transcript-level import — set txOut = TRUE so it does NOT summarize to gene level
txi_SR <- tximport(files, type = "salmon", txOut = TRUE)

tx_counts_SR <- txi_SR$counts
rownames(tx_counts_SR) <- gsub("\\..*", "", rownames(tx_counts_SR))  # strip version suffix

# DTE for SR

library(edgeR)

y_tx_SR <- DGEList(counts = tx_counts_SR, samples = LR_samples)  # same metadata structure as before
keep <- filterByExpr(y_tx_SR)
y_tx_SR <- y_tx_SR[keep, , keep.lib.sizes = FALSE]
y_tx_SR <- normLibSizes(y_tx_SR)

design <- model.matrix(~ Donor + Treatment)
y_tx_SR <- estimateDisp(y_tx_SR, design)
fit_tx_SR <- glmQLFit(y_tx_SR, design)
lrt_tx_SR <- glmQLFTest(fit_tx_SR, coef = "Treatment4Gy")

edger_tx_SR_res <- topTags(lrt_tx_SR, n = Inf)$table

# check the two IL32 transcripts specifically
il32_tx <- c("ENST00000325568", "ENST00000440815")
edger_tx_SR_res[rownames(edger_tx_SR_res) %in% il32_tx, ]

# DTU

tx2gene_SR <- transcript_to_gene[transcript_to_gene$ensembl_transcript_id %in% rownames(tx_counts_SR), ]

mat_SR <- tx_counts_SR[rownames(tx_counts_SR) %in% tx2gene_SR$ensembl_transcript_id, ]
gene_df_SR <- data.frame(
  feature_id = rownames(mat_SR),
  gene_id = tx2gene_SR$ensembl_gene_id[match(rownames(mat_SR), tx2gene_SR$ensembl_transcript_id)]
)

y_SR <- DGEList(counts = mat_SR, samples = LR_samples, genes = gene_df_SR)
keep <- filterByExpr(y_SR, group = LR_metadata$Dose)
y_SR <- y_SR[keep, , keep.lib.sizes = FALSE]
y_SR <- normLibSizes(y_SR)

design <- model.matrix(~ Donors + group, data = y_SR$samples)
fit_SR <- glmQLFit(y_SR, design)

ex_SR <- diffSplice(fit_SR, coef = 4, geneid = "gene_id", exonid = "feature_id", robust = TRUE)

# check WDR74 and ITM2B specifically
wdr74_id <- "ENSG00000105372"   # you already have this from your LR DTU code
itm2b_id <- tx2gene_SR$ensembl_gene_id[tx2gene_SR$external_gene_name == "ITM2B"][1]

ts_gene_SR <- topSplice(ex_SR, number = Inf)

# transcript-level result — mirrors your LR "ts.tx"
ts_tx_SR <- topSplice(ex_SR, test = "t", number = Inf)

# stageR correction

# gene-level screening p-values (Simes), same structure as your LR "pScreen"
pScreen_SR <- ts_gene_SR$P.Value
names(pScreen_SR) <- ts_gene_SR$gene_id

# transcript-level confirmation p-values, same structure as your LR "pConfirmation"
pConfirmation_SR <- matrix(ts_tx_SR$P.Value, ncol = 1)
rownames(pConfirmation_SR) <- ts_tx_SR$feature_id

tx2gene_SR_map <- ts_tx_SR[, c("feature_id", "gene_id")]

stageRObj_SR <- stageRTx(
  pScreen = pScreen_SR,
  pConfirmation = pConfirmation_SR,
  pScreenAdjusted = FALSE,
  tx2gene = tx2gene_SR_map
)

stageRObj_SR <- stageWiseAdjustment(
  object = stageRObj_SR,
  method = "dtu",
  alpha = 0.05
)

suppressWarnings({
  ds.padj_SR <- getAdjustedPValues(stageRObj_SR, order = TRUE, onlySignificantGenes = TRUE)
})

# now check WDR74 and ITM2B against the stageR-adjusted result
ds.padj_SR[ds.padj_SR$geneID %in% c(wdr74_id, itm2b_id), ]

```

```{r}
# gene level
# the exact gene set edgeR tested, from your existing gene-level DEA block
edger_tested_genes <- rownames(y_DEA)   # already filtered + kept.lib.sizes=FALSE applied

# subset dds_LR to the same gene set — no separate filtering logic, just match edgeR's set exactly
dds_LR_matched <- dds_LR[rownames(dds_LR) %in% edger_tested_genes, ]

nrow(dds_LR_matched)   # sanity check — should equal length(edger_tested_genes) (~13,020)

# run DESeq2 on the matched feature set, same design as before
dds_LR_matched <- DESeq(dds_LR_matched)
res_matched <- results(dds_LR_matched, name = "Dose_4Gy_vs_0Gy")  # confirm exact coefficient name via resultsNames(dds_LR_matched)

# significant genes at FDR <= 0.05
sig_deseq2_matched <- res_matched[!is.na(res_matched$padj) & res_matched$padj <= 0.05, ]
nrow(sig_deseq2_matched)

# p-value histogram — the one both reviewers scrutinized
hist(res_matched$pvalue)

# logFC correlation with edgeR on the same matched gene set
common_genes <- intersect(rownames(sig_genes_edger), rownames(res_matched))  # or better: full tested set, not just edgeR's sig genes
edger_logfc <- edger_gene_res$table[rownames(res_matched), "logFC"]
cor(edger_logfc, res_matched$log2FoldChange, use = "complete.obs")

# transcript level

# the exact transcript set edgeR tested (from your transcript-level DEA block, y_DEA_tx after filterByExpr)

strip_version_safe <- function(ids) {
  ifelse(grepl("^ENST", ids),
         gsub("\\..*", "", ids),   # strip version suffix only for annotated transcripts
         ids)                       # leave novel "transcriptN.chrX.category" IDs fully intact
}

dds_tx_raw <- DESeqDataSetFromMatrix(countData = tx_mat, colData = LR_metadata, design = ~ Dose + Donors)
rownames(dds_tx_raw) <- strip_version_safe(rownames(dds_tx_raw))
sum(duplicated(rownames(dds_tx_raw)))   # should now be 0


edger_tested_tx <- strip_version_safe(rownames(y_DEA_tx))


dds_tx_matched <- dds_tx_raw[rownames(dds_tx_raw) %in% edger_tested_tx, ]
nrow(dds_tx_matched)


dds_tx_matched <- DESeq(dds_tx_matched)
res_tx_matched <- results(dds_tx_matched, name = "Dose_4Gy_vs_0Gy")
sig_deseq2_tx_matched <- res_tx_matched[!is.na(res_tx_matched$padj) & res_tx_matched$padj <= 0.05, ]
nrow(sig_deseq2_tx_matched)

# also re-check the edgeR side for the same duplicate issue before finalizing the transcript-level DEG count
sum(duplicated(rownames(edger_Tx_res$table)))

hist(res_tx_matched$pvalue)

edger_tx_logfc <- edger_Tx_res$table[rownames(res_tx_matched), "logFC"]
cor(edger_tx_logfc, res_tx_matched$log2FoldChange, use = "complete.obs")


# venn diagram

library(VennDiagram)

# ---- Gene level ----
edger_sig_genes <- rownames(sig_genes_edger)                      # 183, already computed earlier
deseq2_sig_genes <- rownames(sig_deseq2_matched)                  # 120, from the matched rerun

length(edger_sig_genes)
length(deseq2_sig_genes)
length(intersect(edger_sig_genes, deseq2_sig_genes))               # overlap
length(setdiff(edger_sig_genes, deseq2_sig_genes))                 # edgeR-only
length(setdiff(deseq2_sig_genes, edger_sig_genes))                 # DESeq2-only

venn_gene <- venn.diagram(
  x = list(edgeR = edger_sig_genes, DESeq2 = deseq2_sig_genes),
  filename = NULL,
  fill = c("#66c2a5", "#8da0cb"),
  alpha = 0.6,
  cex = 1.4,
  cat.cex = 1.4,
  main = "Gene-level DEGs (FDR \u2264 0.05)"
)
grid::grid.newpage()
grid::grid.draw(venn_gene)

# ---- Transcript level ----
edger_sig_tx <- strip_version_safe(rownames(sig_tx))                                  # from your existing edgeR transcript-level block
deseq2_sig_tx <- rownames(sig_deseq2_tx_matched)                  # 71, from the matched rerun

length(edger_sig_tx)
length(deseq2_sig_tx)
length(intersect(edger_sig_tx, deseq2_sig_tx))
length(setdiff(edger_sig_tx, deseq2_sig_tx))
length(setdiff(deseq2_sig_tx, edger_sig_tx))

venn_tx <- venn.diagram(
  x = list(edgeR = edger_sig_tx, DESeq2 = deseq2_sig_tx),
  filename = NULL,
  fill = c("#66c2a5", "#8da0cb"),
  alpha = 0.6,
  cex = 1.4,
  cat.cex = 1.4,
  main = "Transcript-level DETs (FDR \u2264 0.05)"
)
grid::grid.newpage()
grid::grid.draw(venn_tx)

# save both cleanly if you want them as standalone figure files
ggsave("Fig2C_gene_venn.png", plot = venn_gene, width = 6, height = 6, dpi = 400)
ggsave("Fig2C_tx_venn.png", plot = venn_tx, width = 6, height = 6, dpi = 400)
```

```{r}
## plot lfc vs lfc for genes and transcripts

lfc_df <- data.frame(
  edgeR = edger_tx_logfc,
  DESeq2 = res_tx_matched$log2FoldChange
)

lfc_df <- lfc_df[complete.cases(lfc_df), ]

# Correlation
cor_test <- cor.test(lfc_df$edgeR, lfc_df$DESeq2, method = "pearson")

ggplot(lfc_df, aes(x = edgeR, y = DESeq2)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    hjust = 1.1,
    vjust = -0.5,
    label = sprintf(
      "Pearson r = %.3f\nP.value = %.2e\nN = %d",
      cor_test$estimate,
      cor_test$p.value,
      nrow(lfc_df)
    ),
    size = 4
  ) +
  labs(
    x = "edgeR transcript log2FC",
    y = "DESeq2 transcript log2FC",
    title = "Comparison of Gene-level log2 Fold Changes: DESeq2 vs EdgeR"
  ) +
  theme_classic(base_size = 14) +ggpt



edger_logfc <- edger_gene_res$table[rownames(res_matched), "logFC"]

gene_lfc_df <- data.frame(
  edgeR = edger_logfc,
  DESeq2 = res_matched$log2FoldChange
)

# Remove missing/infinite values
gene_lfc_df <- gene_lfc_df[
  is.finite(gene_lfc_df$edgeR) &
  is.finite(gene_lfc_df$DESeq2),
]

# Pearson correlation
gene_cor <- cor.test(
  gene_lfc_df$edgeR,
  gene_lfc_df$DESeq2,
  method = "pearson"
)

gene_cor

ggplot(gene_lfc_df, aes(x = edgeR, y = DESeq2)) +
  geom_point(alpha = 0.6, size = 2) +
  
  # Perfect agreement: y = x
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    hjust = 1.1,
    vjust = -0.5,
    label = sprintf(
      "Pearson r = %.3f\nP.value = %.2e\nN = %d",
      gene_cor$estimate,
      gene_cor$p.value,
      nrow(gene_lfc_df)
    ),
    size = 4
  ) +
  
  labs(
    x = "edgeR gene log2FC",
    y = "DESeq2 gene log2FC",
    title = "Comparison of Gene-level log2 Fold Changes: DESeq2 vs EdgeR"
  ) +
  
  theme_classic(base_size = 14) + ggpt

```

## R2
```{r}
# confirm you have the list of 26 novel transcript IDs, in the same "transcriptN.chrX.category" format as the raw IsoQuant/tmap qry_id column
novel_26_ids <- rownames(sig_novel_tx)   # adjust if your final 26 list lives elsewhere (e.g. after TransDecoder/CPC2/CPAT filtering)
length(novel_26_ids)   # sanity check — should be 26

# check the tmap qry_id column format matches
head(tmap$qry_id)

# filter tmap to just these 26
tmap_26 <- tmap[tmap$qry_id %in% novel_26_ids, ]
nrow(tmap_26)   # should be 26 — if fewer, some IDs didn't match (format mismatch, check strip_version_safe-type issues again)

table(tmap_26$class_code)

# full per-transcript table for the supplement
tmap_26[, c("qry_id", "ref_gene_id", "class_code", "ref_id", "len", "ref_len")]


# DGE on the SR
dds_SR <- readRDS("~/Phybion_Pilot_LR/dds_SR.RDS")

dds_SR <- DESeq(dds_SR)
resultsNames(dds_SR)   # confirm exact coefficient name
res_SR <- results(dds_SR, name = "Dose_4Gy_vs_0Gy")   # adjust to actual coefficient name

sig_genes_SR <- res_SR[!is.na(res_SR$padj) & res_SR$padj <= 0.05, ]
nrow(sig_genes_SR)


sr_deg_ids <- rownames(sig_genes_SR)   # your SR significant-gene Ensembl IDs, stripped of version suffix
lr_deg_ids <- rownames(sig_genes_edger)    # your 183 LR DEGs, already Ensembl IDs

length(sr_deg_ids); length(lr_deg_ids)
length(intersect(sr_deg_ids, lr_deg_ids))          # shared
length(setdiff(lr_deg_ids, sr_deg_ids))            # LR-only
length(setdiff(sr_deg_ids, lr_deg_ids))  
```

```{r}
# pull all 34 genes' LR results, regardless of significance (edger_gene_res is your full table, not just sig hits)
signature_34 <-c("ENSG00000198522",
"ENSG00000122687",
"ENSG00000123689",
"ENSG00000145365",
"ENSG00000160803",
"ENSG00000141682",
"ENSG00000120278",
"ENSG00000228526",
"ENSG00000157557",
"ENSG00000105426",
"ENSG00000102879",
"ENSG00000172716",
"ENSG00000249437",
"ENSG00000167483",
"ENSG00000197694",
"ENSG00000068028",
"ENSG00000160469",
"ENSG00000101773",
"ENSG00000172115",
"ENSG00000065183",
"ENSG00000149554",
"ENSG00000119917",
"ENSG00000277925",
"ENSG00000185515",
"ENSG00000099203",
"ENSG00000234127",
"ENSG00000106052",
"ENSG00000173559",
"ENSG00000172613",
"ENSG00000035664",
"ENSG00000204569",
"ENSG00000160953",
"ENSG00000163945",
"ENSG00000197774")   #  34 IDs here

lr_signature_res <- edger_gene_res$table[rownames(edger_gene_res$table) %in% signature_34, ]
lr_signature_res$SYMBOL <- mapIds(org.Hs.eg.db, keys = rownames(lr_signature_res),
                                   column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")

dim(lr_signature_res)


res_SR<- as.data.frame(res_SR)

sr_signature_values <- res_SR[rownames(res_SR) %in% signature_34, ]


dim(sr_signature_values)

# merge with the original SR logFC/FDR values you'll supply
comparison_table <- merge(
    lr_signature_res,
    sr_signature_values,
    by = "row.names"
)

comparison_table <- comparison_table %>%
  rename(
    logFC_LR_edgeR = logFC,
    logCPM_LR_edgeR = logCPM,
    F_LR_edgeR = F,
    PValue_LR_edgeR = PValue,
    FDR_LR_edgeR = FDR,
    
    baseMean_SR_DESeq2 = baseMean,
    log2FoldChange_SR_DESeq2 = log2FoldChange,
    lfcSE_SR_DESeq2 = lfcSE,
    stat_SR_DESeq2 = stat,
    pvalue_SR_DESeq2 = pvalue,
    padj_SR_DESeq2 = padj
  )

write_xlsx(comparison_table, "comparison_34genes.xlsx")
# 1. directional concordance — simple sign test
concordant <- sign(comparison_table$logFC) == sign(comparison_table$log2FoldChange)
sum(concordant); length(concordant)
binom.test(sum(concordant), length(concordant), p = 0.5)   # is concordance better than chance?

# 2. set-level test — treat the 34-gene signature as a single gene set, tested in the LR data
library(limma)

keep <- rowSums(y_DEA$counts > 0) > 0

y_DEA <- estimateDisp(y_DEA, design)

y_mroast <- y_DEA[keep, , keep.lib.sizes = FALSE]

idx <- ids2indices(
    list(radioresponsive_34 = signature_34),
    identifiers = rownames(y_mroast)
)

# idx <- ids2indices(list(radioresponsive_34 = signature_34), identifiers = rownames(y_DEA))

design_gene <- model.matrix(~ Donor + Treatment, data = data.frame(Donor = factor(LR_metadata$Donors), Treatment = factor(LR_metadata$Dose)))

mroast(y_mroast, index = idx, design = design, contrast = "Treatment4Gy")
# or: 
camera(y_mroast, index = idx, design = design_gene, contrast = "Treatment4Gy")
```

```{r}
#Comment 7

library(data.table)
head(read_assign)
read_assign <- fread("Isoquant/OUT/OUT.read_assignments.tsv.gz", skip = 1)
colnames(read_assign) <- c("read_id", "chr", "strand", "isoform_id", "gene_id",
                            "assignment_type", "assignment_events", "exons", "additional_info")

table(read_assign$assignment_type)  # or whatever the classification column is named — check colnames first


library(rtracklayer)
isoquant_gtf <- import("./Isoquant/OUT/OUT.transcript_models.gtf")

novel_26_gr <- isoquant_gtf[isoquant_gtf$transcript_id %in% novel_26_ids]
export(novel_26_gr, "novel_26_transcripts.gtf", format = "gtf")

# venn

library(VennDiagram)

# Set membership, derived from the coding-potential table
cpc2_coding <- c("transcript61117.chr1.nnic", "transcript23187.chr5.nnic", "transcript13760.chr9.nnic",
                 "transcript21664.chr9.nnic", "transcript15213.chr10.nnic", "transcript36523.chr12.nnic",
                 "transcript13826.chr15.nnic", "transcript13458.chr22.nnic")

cpat_coding <- c("transcript61117.chr1.nnic", "transcript30521.chr2.nnic", "transcript23187.chr5.nnic",
                 "transcript21609.chr6.nnic", "transcript37812.chr6.nnic", "transcript13760.chr9.nnic",
                 "transcript21664.chr9.nnic", "transcript15213.chr10.nnic", "transcript36523.chr12.nnic",
                 "transcript13826.chr15.nnic", "transcript13458.chr22.nnic")

transdecoder_any_orf <- c("transcript13458.chr22.nnic", "transcript13760.chr9.nnic", "transcript13826.chr15.nnic",
                          "transcript15213.chr10.nnic", "transcript21609.chr6.nnic", "transcript21664.chr9.nnic",
                          "transcript23187.chr5.nnic", "transcript25470.chr1.nic", "transcript30521.chr2.nnic",
                          "transcript36523.chr12.nnic", "transcript61117.chr1.nnic")

venn_coding <- venn.diagram(
  x = list(CPC2 = cpc2_coding, CPAT = cpat_coding, TransDecoder = transdecoder_any_orf),
  filename = NULL,
  fill = c("#1B9E77", "#D95F02", "#7570B3"),
  alpha = 0.6,
  cex = 1.4,
  cat.cex = 1.3,
  main = "Coding-potential predictions across 26 novel/unannotated transcripts"
)
grid::grid.newpage()
grid::grid.draw(venn_coding)

ggsave("Fig5A_coding_potential_venn.png", plot = venn_coding, width = 6, height = 6, dpi = 400)
```

### supsampling

```{r}
## =============================================================================
## Read-depth saturation analysis (subSeq) for long-read gene-level DGE
## Addresses reviewer comments on gene detection sensitivity vs. sequencing depth
## =============================================================================

library(subSeq)
library(edgeR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(org.Hs.eg.db)

## -----------------------------------------------------------------------------
## 1. Inputs
## -----------------------------------------------------------------------------

# Filtered gene count matrix (filterByExpr already applied on full-depth data,
# kept fixed across all subsampling depths so the gene universe is consistent)
counts_filtered <- y_DEA$counts

Donor     <- factor(LR_metadata$Donors)
Treatment <- factor(LR_metadata$Dose)

## -----------------------------------------------------------------------------
## 2. Custom edgeR QLF handler (mirrors the original DGE pipeline exactly,
##    including donor-blocking; the built-in subSeq "edgeR" handler does not
##    support this design and must not be used here)
## -----------------------------------------------------------------------------

my_edgeR_qlf <- function(count.matrix, treatment, donor, coef_name, ...) {

  design <- model.matrix(~ donor + treatment)

  if (!coef_name %in% colnames(design)) {
    stop(
      "Coefficient '", coef_name, "' not found. Available coefficients: ",
      paste(colnames(design), collapse = ", ")
    )
  }

  y   <- DGEList(counts = count.matrix)
  y   <- normLibSizes(y)
  y   <- estimateDisp(y, design)
  fit <- glmQLFit(y, design)
  lrt <- glmQLFTest(fit, coef = coef_name)
  res <- topTags(lrt, n = Inf, sort.by = "none")$table

  # subSeq requires exactly one row per gene in the input matrix, in order
  out <- data.frame(
    pvalue      = rep(1, nrow(count.matrix)),
    coefficient = rep(0, nrow(count.matrix)),
    row.names   = rownames(count.matrix)
  )
  out[rownames(res), "pvalue"]      <- res$PValue
  out[rownames(res), "coefficient"] <- res$logFC
  out
}

## -----------------------------------------------------------------------------
## 3. Run the saturation analysis
## -----------------------------------------------------------------------------

proportions <- c(0.05, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1.00)

set.seed(42)
ss <- subsample(
  counts_filtered,
  proportions,
  method       = my_edgeR_qlf,
  replications = 20,
  treatment    = Treatment,
  donor        = Donor,
  coef_name    = "treatment4Gy"
)

ss_df <- as.data.frame(ss)

## -----------------------------------------------------------------------------
## 4. Full-depth (oracle) DEG list, BH-corrected to match the original DGE pipeline
##    (do NOT use the subSeq-internal $qvalue column -- q-value gives
##    a different and more permissive correction than the BH/topTags FDR used
##    to define the manuscript's DEG list)
## -----------------------------------------------------------------------------

oracle <- ss_df %>%
  filter(proportion == 1) %>%
  mutate(FDR_BH = p.adjust(pvalue, method = "BH"))

degs_full <- oracle %>% filter(FDR_BH <= 0.05) %>% pull(ID)
length(degs_full)   # should match your original edgeR DEG count (e.g. 183)

## -----------------------------------------------------------------------------
## 5. Per-gene detection frequency at each subsampled depth
##    (BH applied within each proportion x replication group, since each
##    represents an independent differential expression run)
## -----------------------------------------------------------------------------

detection_freq <- ss_df %>%
  filter(proportion < 1) %>%
  group_by(proportion, replication) %>%
  mutate(FDR_BH = p.adjust(pvalue, method = "BH")) %>%
  ungroup() %>%
  filter(ID %in% degs_full) %>%
  group_by(proportion, ID) %>%
  summarise(
    times_sig      = sum(FDR_BH <= 0.05),
    n_reps         = n(),
    detection_rate = times_sig / n_reps,
    .groups = "drop"
  )

## -----------------------------------------------------------------------------
## 6. Depth-level summary (overall recovery curve)
## -----------------------------------------------------------------------------

depth_summary <- detection_freq %>%
  group_by(proportion) %>%
  summarise(mean_detection = mean(detection_rate), .groups = "drop") %>%
  bind_rows(tibble(proportion = 1, mean_detection = 1))

print(depth_summary)

## -----------------------------------------------------------------------------
## 7. Supplementary table
## -----------------------------------------------------------------------------

oracle_stats <- oracle %>%
  filter(ID %in% degs_full) %>%
  dplyr::select(ID, oracle_logFC = coefficient, oracle_FDR = FDR_BH)

detection_wide <- detection_freq %>%
  mutate(proportion_label = paste0(proportion * 100, "pct")) %>%
  dplyr::select(ID, proportion_label, detection_rate) %>%
  pivot_wider(names_from = proportion_label, values_from = detection_rate,
              names_prefix = "detect_")

supp_table <- oracle_stats %>%
  left_join(detection_wide, by = "ID") %>%
  mutate(
    mean_detection = rowMeans(dplyr::select(., starts_with("detect_")), na.rm = TRUE),
    category = case_when(
      mean_detection >= 0.9 ~ "Robust",
      mean_detection >= 0.5 ~ "Intermediate",
      TRUE                  ~ "Depth-sensitive"
    ),
    effect_tertile = ntile(abs(oracle_logFC), 3),
    effect_tertile = factor(effect_tertile,
                             labels = c("Low |logFC|", "Mid |logFC|", "High |logFC|"))
  ) %>%
  arrange(desc(mean_detection))

supp_table$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys      = supp_table$ID,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

write.csv(supp_table, "./SupplementaryTable_DetectionSaturation.csv", row.names = FALSE)

## -----------------------------------------------------------------------------
## 8. Supplementary figure
## -----------------------------------------------------------------------------

heatmap_df <- detection_freq %>%
  mutate(ID = factor(ID, levels = supp_table$ID))   # ordered by robustness

p_heatmap <- ggplot(heatmap_df, aes(x = factor(proportion * 100), y = ID, fill = detection_rate)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "white", high = "#08519c", limits = c(0, 1),
                       name = "Detection\nfrequency") +
  labs(x = "Subsampled depth (% of full)", y = NULL, title = "A") +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

trajectory_df <- detection_freq %>%
  left_join(supp_table %>% dplyr::select(ID, effect_tertile), by = "ID") %>%
  group_by(proportion, effect_tertile) %>%
  summarise(mean_detection = mean(detection_rate),
            se = sd(detection_rate) / sqrt(n()), .groups = "drop")

p_trajectory <- ggplot(trajectory_df, aes(x = proportion, y = mean_detection, color = effect_tertile)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_detection - se, ymax = mean_detection + se), width = 0.01) +
  scale_x_continuous(breaks = proportions[proportions < 1],
                      labels = function(x) paste0(x * 100, "%")) +
  scale_color_manual(values = c("#fcae91", "#fb6a4a", "#a50f15")) +
  labs(x = "Subsampled depth", y = "Mean detection frequency",
       color = "Effect size\ntertile", title = "B") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right")

p_final <- p_heatmap + p_trajectory + plot_layout(widths = c(1, 1.3))
ggsave("./SupplementaryFigure_DetectionSaturation.pdf",
       p_final, width = 9, height = 6)
```


```{r}
library(DoReMiTra)
library(SummarizedExperiment)
library(ggplot2)

# ============================================================
# 1. Retrieve datasets
# ============================================================


DoReMiTra::search_DoReMiTra_datasets(
  radiation_type = "x-ray",
  exp_setting = "exvivo"
)


se1 <- get_DoReMiTra_data(
  "SE_Broustas_2017_ExVivo_GSE90909_GPL13497"
)

se2 <- get_DoReMiTra_data(
  "SE_Ghandhi_2015_ExVivo_GSE65292_GPL13497"
)


se3 <- get_DoReMiTra_data(
  "SE_Salah_2025_ExVivo"
)


# ============================================================
# 2. ACTA2 identifiers
# ============================================================

probe_id <- "A_23_P150053"
ensembl_id <- "ENSG00000107796"


# ============================================================
# 3. Split Salah dataset by time point
# ============================================================

se3_2h <- se3[, se3$Time_point == "2hrs"]
se3_6h <- se3[, se3$Time_point == "6hrs"]


# ============================================================
# 4. Plotting function
# ============================================================

plot_gene_response <- function(se, feature_id, dataset_name) {

  df <- data.frame(
    Dose = as.character(colData(se)$Dose),
    expression = as.numeric(assay(se)[feature_id, ])
  )

  # Remove missing doses
  df <- df[!is.na(df$Dose), ]

  # Extract numeric dose for ordering
  df$dose_numeric <- as.numeric(
    gsub(" Gy|Gy", "", df$Dose)
  )

  # Order doses numerically
  df$Dose <- factor(
    df$Dose,
    levels = unique(df$Dose[order(df$dose_numeric)])
  )

  ggplot(df, aes(x = Dose, y = expression, fill = Dose)) +
    geom_boxplot(
      width = 0.6,
      outlier.shape = NA
    ) +
    geom_jitter(
      width = 0.12,
      size = 2.5
    ) +
    theme_classic() +
    labs(
      title = dataset_name,
      subtitle = "ACTA2",
      x = "Dose",
      y = "Expression"
    ) +
    guides(fill = "none")
}


# ============================================================
# 5. Generate individual plots
# ============================================================

p1 <- plot_gene_response(
  se1,
  probe_id,
  "Broustas et al. 2017 — 24h"
)

p2 <- plot_gene_response(
  se2,
  probe_id,
  "Ghandhi et al. 2015 — 24h"
)


p3_2h <- plot_gene_response(
  se3_2h,
  ensembl_id,
  "Salah et al. 2025 — 2 h"
)

p3_6h <- plot_gene_response(
  se3_6h,
  ensembl_id,
  "Salah et al. 2025 — 6 h"
)


# ============================================================
# 6. Display individually
# ============================================================

p1
p2
p3_6h
p3_2h # the response present from 6h
```

```{r}
library(dplyr)
library(readr)
library(writexl)

# ============================================================
# 1. SQANTI3 — classification + filter results, cleaned
# ============================================================

# LR-only run
sqanti_class_lr <- read_tsv("SQANTI3/sqanti3_out_full/sqanti3_novel26_full_classification.txt")
sqanti_filtered_lr <- read_tsv("SQANTI3/sqanti3_out_full/sqanti3_novel26_full_filtered_RulesFilter_classification.txt")

# With short-read support
sqanti_class_sr <- read_tsv("SQANTI3/sqanti3_out_withSR2/sqanti3_novel26_withSR_classification.txt")
sqanti_filtered_sr <- read_tsv("SQANTI3/sqanti3_out_withSR2/sqanti3_novel26_withSR_filtered_RulesFilter_classification.txt")

# Keep just the columns that matter for a clean summary table
sqanti_clean <- sqanti_filtered_sr %>%
  dplyr::select(isoform, structural_category, subcategory, associated_gene,
         all_canonical, min_sample_cov, min_cov, min_cov_pos, RTS_stage,
         coding, CDS_type, psauron_score, filter_result) %>%
  arrange(desc(filter_result == "Isoform"), isoform)

write_xlsx(sqanti_clean, "SQANTI3_novel26_summary_withSR.xlsx")

# ============================================================
# 2. SQANTI3 — junctions, cleaned (per-junction detail)
# ============================================================

sqanti_junc <- read_tsv("SQANTI3/sqanti3_out_withSR2/sqanti3_novel26_withSR_junctions.txt") %>%
  select(isoform, junction_number, chrom, strand,
         genomic_start_coord, genomic_end_coord,
         junction_category, canonical, splice_site,
         RTS_junction, bite_junction,
         total_coverage_unique, total_coverage_multi)

write_xlsx(sqanti_junc, "SQANTI3_novel26_junctions_cleaned.xlsx")

# ============================================================
# 3. BLASTp — parse your outfmt 6 tabular results
# ============================================================

blast_cols <- c("qseqid", "sseqid", "pident", "length", "E-value",
                 "Bit score", "qcovs", "stitle")

blastp <- read_excel("Additional_manuscript_Supp/untitled folder/Additional_file14_BLASTp_results.xlsx",col_names = blast_cols )
# (adjust filename/path to wherever your real blastp -outfmt 6 output lives)

blastp_clean <- blastp %>%
  mutate(gene = case_when(
    grepl("13458", qseqid) ~ "APOBEC3C",
    grepl("13760", qseqid) ~ "TNFSF8",
    grepl("13826", qseqid) ~ "FBXO22",
    grepl("15213", qseqid) ~ "ACTA2",
    grepl("21664", qseqid) ~ "PHPT1",
    grepl("23187", qseqid) ~ "CCNG1",
    grepl("36523", qseqid) ~ "PRKAB1",
    TRUE ~ NA_character_
  )) %>%
  select(gene, qseqid, sseqid, pident, length, evalue, bitscore, qcovs, stitle) %>%
  arrange(gene, evalue)

write_xlsx(blastp_clean, "BLASTp_results_cleaned.xlsx")

# ============================================================
# 4. InterProScan — parse the raw TSV (no header by default)
# ============================================================

ipr_cols <- c("protein_id", "md5", "length", "analysis", "signature_acc",
              "signature_desc", "start", "end", "score", "status", "date",
              "interpro_acc", "interpro_desc", "go_terms", "pathways")

interpro <- read_tsv("iprscan5-I20260721-114307-0199-33477862-p1m.tsv",
                      col_names = ipr_cols, na = "-")

# drop MobiDBLite (disorder prediction, not a real structural domain)
interpro_domains <- interpro %>%
  filter(analysis == "Pfam") %>%
  dplyr::select(protein_id, length, analysis, signature_acc, signature_desc,
         start, end, score, interpro_acc, interpro_desc)

write_xlsx(interpro_domains, "InterProScan_domains_cleaned_pfam.xlsx")

# a compact one-row-per-protein summary: which domains does each candidate/reference carry
interpro_summary <- interpro_domains %>%
  group_by(protein_id) %>%
  summarise(domains = paste(unique(na.omit(interpro_desc)), collapse = "; "),
            n_domain_hits = n())

write_xlsx(interpro_summary, "InterProScan_domain_summary.xlsx")
```

```{r}
ipr <- read_tsv("iprscan5-I20260721-114307-0199-33477862-p1m.tsv", col_names = FALSE, na = "-")

colnames(ipr)[1:9] <- c("protein_id", "md5", "length", "analysis", "signature_acc",
                        "signature_desc", "start", "end", "score")

# keep only real domain-calling analyses (drop MobiDBLite disorder-only predictions,
# which aren't structural domains)
ipr_domains <- ipr %>% filter(analysis == "Pfam")

# order proteins so each candidate sits directly above its reference
# protein_order <- c(
#     "APOBEC3C_candidate_transcript13458_p1", "APOBEC3C_reference",
#     "TNFSF8_candidate_transcript13760_p1", "TNFSF8_reference",
#     "FBXO22_candidate_transcript13826_p1", "FBXO22_candidate_transcript13826_p2", "FBXO22_reference",
#     "ACTA2_candidate_transcript15213_p1", "ACTA2_reference",
#     "CCNG1_candidate_transcript23187_p1", "CCNG1_reference",
#     "PHPT1_candidate_transcript21664_p1", "PHPT1_reference",
#     "PRKAB1_candidate_transcript36523_p1", "PRKAB1_reference"
# )

protein_order <- c(
    "FBXO22_candidate_transcript13826_p1", "FBXO22_candidate_transcript13826_p2", "FBXO22_reference",
    "PHPT1_candidate_transcript21664_p1", "PHPT1_reference"
)

ipr_domains <- ipr_domains %>%
    filter(protein_id %in% protein_order) %>%
    mutate(protein_id = factor(protein_id, levels = rev(protein_order)))

# protein backbone lengths (one row per protein, for the background line)
backbones <- ipr_domains %>% distinct(protein_id, length)

ggplot() +
    geom_segment(data = backbones, aes(x = 0, xend = length, y = protein_id, yend = protein_id),
                 color = "grey70", linewidth = 1.2) +
    geom_rect(data = ipr_domains,
              aes(xmin = start, xmax = end,
                  ymin = as.numeric(protein_id) - 0.25, ymax = as.numeric(protein_id) + 0.25,
                  fill = signature_desc)) +
    labs(x = "Amino acid position", y = NULL, fill = "Domain") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", legend.text = element_text(size = 8)) +
    guides(fill = guide_legend(ncol = 2))

```

```{r}
# CONFIG - edit these paths for your setup
# =============================================================================
 
classification_path <-"sqanti_new_test/sqanti3_out_subSR/sqanti3_novel26_subSR_filtered_RulesFilter_classification.txt"
diagnostics_path     <- "sqanti_new_test/novel_junction_diagnostics.tsv"
crosscheck_path      <- "sqanti_new_test/isoquant_crosscheck.tsv"
 
output_xlsx <- "Supplementary_SQANTI3_diagnostics_combined.xlsx"
 
keep_filter_result <- "Isoform"   # only report filter-passed transcripts
 
# =============================================================================
# 1. Load and trim SQANTI3 classification
# =============================================================================
 
classification_raw <- read_tsv(classification_path, show_col_types = FALSE)
 
classification <- classification_raw %>%
  filter(filter_result == keep_filter_result) %>%
  dplyr::select(any_of(c(
    "isoform", "associated_gene", "structural_category", "subcategory",
    "min_sample_cov", "min_cov", "min_cov_pos", "sd_cov",
    "all_canonical", "bite", "coding", "filter_result"
  ))) %>%
  arrange(isoform)
 
# =============================================================================
# 2. Load junction-level diagnostics (SR + LR, per junction)
# =============================================================================
 
diagnostics <- read_tsv(diagnostics_path, show_col_types = FALSE) %>%
  dplyr::select(any_of(c(
    "transcript", "junction_number", "chrom", "start", "end", "strand",
    "SR_exact_total_reads", "SR_exact_n_samples_supporting",
    "SR_locus_active", "SR_locus_max_reads",
    "LR_n_bams_supporting", "LR_n_bams_total", "LR_total_reads",
    "verdict"
  )))
 
# =============================================================================
# 3. Load IsoQuant crosscheck and summarize per transcript
#    (input is one row per transcript x sample; collapse across the 6 libraries)
# =============================================================================
 
crosscheck_raw <- read_tsv(crosscheck_path, show_col_types = FALSE)
 
crosscheck_summary <- crosscheck_raw %>%
  group_by(transcript) %>%
  summarise(
    isoquant_assigned_reads_total     = sum(isoquant_assigned_reads, na.rm = TRUE),
    cigar_confirmed_all_junctions_total  = sum(cigar_confirmed_all_junctions, na.rm = TRUE),
    cigar_confirmed_some_junctions_total = sum(cigar_confirmed_some_junctions, na.rm = TRUE),
    n_expected_junctions              = first(n_expected_junctions),
    .groups = "drop"
  ) %>%
  mutate(
    full_confirmation_rate_pct = round(
      100 * cigar_confirmed_all_junctions_total / isoquant_assigned_reads_total, 2
    )
  )
 
# =============================================================================
# 4. Build a single merged, per-transcript summary table
#    (SR/LR figures here are for the transcript's flagged novel/limiting junction,
#     i.e. the row in `diagnostics` matching classification.txt's min_cov_pos)
# =============================================================================
 
#
summary_table <- classification %>%
  left_join(
    diagnostics %>%
      mutate(junction_label = paste0("junction_", junction_number)),
    by = c(
      "isoform" = "transcript",
      "min_cov_pos" = "junction_label"
    )
  ) %>%
  transmute(
    Transcript = isoform,
    Gene_locus = associated_gene,
    Structural_category = structural_category,
    Subcategory = subcategory,
    Limiting_junction_SQANTI3_min_cov_pos = min_cov_pos,
    SQANTI3_isoform_min_cov = min_cov,
    Junction_chrom = chrom,
    Junction_start = start,
    Junction_end = end,
    Novel_junction_SR_reads = SR_exact_total_reads,
    SR_locus_active_nearby = SR_locus_active,
    SR_locus_max_nearby_reads = SR_locus_max_reads,
    Novel_junction_LR_libraries_supporting =
      ifelse(
        is.na(LR_n_bams_supporting),
        NA,
        paste0(LR_n_bams_supporting, "/", LR_n_bams_total)
      ),
    Novel_junction_LR_total_reads = LR_total_reads,
    Novel_junction_verdict = verdict
  ) %>%
  arrange(Transcript)
 
# =============================================================================
# 4. Legend
# =============================================================================
 
legend <- tibble::tribble(
  ~Column, ~Definition_and_source,
  "Transcript / Gene_locus / Structural_category / Subcategory", "SQANTI3 classification.txt output, filter-passed transcripts only.",
  "Limiting_junction_SQANTI3_min_cov_pos", "SQANTI3's own identification of which specific junction has the lowest short-read coverage in the isoform. NOT always junction_1 - varies per transcript - and is the junction all SR/LR figures in this row refer to.",
  "SQANTI3_isoform_min_cov", "SQANTI3's own minimum short-read junction coverage across all junctions of the isoform.",
  "Junction_chrom / Junction_start / Junction_end", "Genomic coordinates of the limiting junction, from independent GTF parsing.",
  "Novel_junction_SR_reads", "Independently re-derived from STAR SJ.out.tab, exact coordinate match, summed across all 6 libraries.",
  "SR_locus_active_nearby / SR_locus_max_nearby_reads", "Whether short-read splicing activity exists anywhere in the surrounding region (+/-5kb), distinguishing genuine absence at the junction from a short-read coverage blind spot in the whole locus.",
  "Novel_junction_LR_libraries_supporting / Novel_junction_LR_total_reads", "Independent long-read CIGAR confirmation (+/-2bp tolerance) at the exact limiting junction coordinates, summed/counted across all 6 libraries.",
  "Novel_junction_verdict", "Plain-language call: SR-CONFIRMED / CANDIDATE NOVEL / INCONCLUSIVE (SR blind spot) / WEAK / LIKELY BOUNDARY-SHIFT.",
  "", "",
  "Tools used", "SQANTI3; STAR (SJ.out.tab); custom Python/pysam script (diagnose_novel_junctions.py) for independent SR/LR/CIGAR validation."
)
 
# =============================================================================
# 5. Write workbook (writexl: simple, no external dependency issues)
# =============================================================================
 
write_xlsx(
  list(Summary_per_transcript = summary_table, Legend = legend),
  path = output_xlsx
)
 
cat("Saved", output_xlsx, "\n")



summary_table <- classification %>%
  left_join(
    diagnostics %>%
      mutate(junction_label = paste0("junction_", junction_number)),
    by = c(
      "isoform" = "transcript",
      "min_cov_pos" = "junction_label"
    )
  ) %>%
  left_join(
    crosscheck_summary,
    by = c("isoform" = "transcript")
  ) %>%
  transmute(
    Transcript = isoform,
    Gene_locus = associated_gene,
    Structural_category = structural_category,
    Subcategory = subcategory,
    Limiting_junction_SQANTI3_min_cov_pos = min_cov_pos,
    SQANTI3_isoform_min_cov = min_cov,
    Novel_junction_SR_reads = SR_exact_total_reads,
    SR_locus_active_nearby = SR_locus_active,
    SR_locus_max_nearby_reads = SR_locus_max_reads,
    Novel_junction_LR_libraries_supporting =
      ifelse(
        is.na(LR_n_bams_supporting),
        NA,
        paste0(LR_n_bams_supporting, "/", LR_n_bams_total)
      ),
    Novel_junction_LR_total_reads = LR_total_reads,
    Novel_junction_verdict = verdict,
    IsoQuant_total_assigned_reads = isoquant_assigned_reads_total,
    IsoQuant_full_isoform_CIGAR_confirmed_reads =
      cigar_confirmed_all_junctions_total,
    IsoQuant_full_confirmation_rate_pct =
      full_confirmation_rate_pct,
    N_expected_junctions = n_expected_junctions
  ) %>%
  arrange(Transcript)
 
# =============================================================================
# 5. Legend
# =============================================================================
 
legend <- tibble::tribble(
  ~Column, ~Definition_and_source,
  "Transcript / Gene_locus / Structural_category / Subcategory", "SQANTI3 classification.txt output, filter-passed transcripts only.",
  "Limiting_junction_SQANTI3_min_cov_pos", "SQANTI3's own identification of which specific junction has the lowest short-read coverage in the isoform. NOT always junction_1 - varies per transcript - and is the junction all SR/LR figures in this row refer to.",
  "SQANTI3_isoform_min_cov", "SQANTI3's own minimum short-read junction coverage across all junctions of the isoform.",
  "Novel_junction_SR_reads", "Independently re-derived from STAR SJ.out.tab, exact coordinate match to the limiting junction, summed across all 6 libraries.",
  "SR_locus_active_nearby / SR_locus_max_nearby_reads", "Whether short-read splicing activity exists anywhere in the surrounding region (+/-5kb), distinguishing genuine absence at the junction from a short-read coverage blind spot in the whole locus.",
  "Novel_junction_LR_libraries_supporting / Novel_junction_LR_total_reads", "Independent long-read CIGAR confirmation (+/-2bp tolerance) at the exact limiting junction coordinates, summed/counted across all 6 libraries.",
  "Novel_junction_verdict", "Plain-language call: SR-CONFIRMED / CANDIDATE NOVEL / INCONCLUSIVE (SR blind spot) / WEAK / LIKELY BOUNDARY-SHIFT.",
  "IsoQuant_total_assigned_reads", "Total long reads IsoQuant assigned to this transcript model, summed across all 6 libraries.",
  "IsoQuant_full_isoform_CIGAR_confirmed_reads / IsoQuant_full_confirmation_rate_pct", "Of the IsoQuant-assigned reads, number/percent independently confirmed by CIGAR to span ALL expected junctions of the transcript in a single read (a stricter, whole-isoform-level check, distinct from the single limiting-junction figures above).",
  "N_expected_junctions", "Total splice junctions in the transcript model, from novel_26_transcripts.gtf.",
  "", "",
  "Tools used", "SQANTI3; STAR (SJ.out.tab); IsoQuant; custom Python/pysam scripts (diagnose_novel_junctions.py, crosscheck_isoquant_assignments.py) for independent SR/LR/CIGAR validation."
)
 
# =============================================================================
# 6. Write workbook
# =============================================================================
 
write_xlsx(
  list(Summary_per_transcript = summary_table, Legend = legend),
  path = output_xlsx
)
 
cat("Saved", output_xlsx, "\n")
```

