---
title: "Phybion_Pilot_LR"
output: html_document
date: "2024-11-26"
---

```{r}
library(knitr)

opts_chunk$set(
  fig.path = paste0('./figures/'),
  fig.align = 'center',
  fig.show = 'asis',
  eval = TRUE,
  fig.width = 10,
  fig.height = 7,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  size = 'small',
  comment = "##",
  echo = TRUE,
  results = 'markup',
  dev = c("png", "pdf")
)
options(replace.assign = TRUE,
        width = 80)

```


# loading the libraries

```{r}
library("readxl")
library("readr")
library("DESeq2")
library("tximeta")
library("SummarizedExperiment")
library("org.Hs.eg.db")
library("macrophage")
library("DESeq2")
library("clusterProfiler")
library("org.Hs.eg.db")
library("enrichplot")
library("mosdef")
library("GeneTonic")
library("IHW")
library("gplots")
library("genefilter")
library("rtracklayer")
library("xtable")
library("GO.db")
library("goseq")
library("GenomicFeatures")
library("ReactomePA")
library("dplyr")
library("NMF")
library("pheatmap")
library("GOplot")
library("knitr")
library("pcaExplorer")
library("DT")
library("ideal")
library("edgeR")
library("hexbin")
library("latticeExtra")
library("vsn")
library("gplots")
library("RColorBrewer")
library("topGO")
library("BIUMmisc")
library("biomaRt")
library("pathview")
library("VennDiagram")
library("GSEABase")
library("msigdbr")
library("ConsensusClusterPlus")
library("visNetwork")
library("igraph")
library("ggraph")
library("decoupleR")
library("OmnipathR")
library("quantiseqr")
library("immunedeconv")
library("tibble")
library("tidyr")
library("ComplexHeatmap")
```

# loading the coldata and count matrix

```{r}
colData<- read_excel("./creating_metadata/coldata_phybion_pilot_LR.xlsx")

colData<- as.data.frame(colData)

count_matrix<- read_tsv("./Isoquant/OUT/OUT.gene_grouped_counts.tsv")

count_matrix<- as.matrix(count_matrix)

rownames(count_matrix)<- count_matrix[,1]

count_matrix<- count_matrix[, -1]

colnames(count_matrix)<- colData$sample_id

rownames(colData)<- colnames(count_matrix)

gene_id<- rownames(count_matrix)

count_matrix <- apply(count_matrix, 2, as.numeric)
rownames(count_matrix)<- gene_id

```

# creating the dds object

```{r}

colData$Dose<- factor(colData$Dose)

colData$Dose <- relevel(colData$Dose, ref = "0Gy")

colData$Sex<- factor(colData$Sex)


dds <- DESeqDataSetFromMatrix(countData = count_matrix,
                              colData = colData,
                              design = ~ Sex+ Dose)
dds

dds <- dds_gencode_to_ensembl(dds)


dds$Dose <- factor(dds$Dose)

dds$Dose <- relevel(dds$Dose, ref = "0Gy")

```

# Variance stablizing transformation

```{r}

vsd_phybion_LR <- vst(dds)


vsd_corrected <- vsd_phybion_LR

assay(vsd_corrected) <- limma::removeBatchEffect(assay(vsd_phybion_LR), batch = vsd_phybion_LR$Donors)

```

# checking read coverage over the samples

```{r}

plot_totcounts(dds, group = "Dose")

plot_totcounts(dds, group = "Sex")

## PCA plot

pca_plot <- plotPCA(vsd_phybion_LR, intgroup = c("Donors", "Dose"), returnData = TRUE)

percentVar <- round(100 * attr(pca_plot, "percentVar"))

ggplot(pca_plot, aes(x = PC1, y = PC2, color = Donors)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17)) + # Adjust the number of shapes if necessary
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()

# clustering samples

sampleDists <- dist(t(assay(vsd_corrected)))

sampleDistMatrix <- as.matrix(sampleDists)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)


# Extract sex from sample names:
sex <- ifelse(grepl("F$", colnames(sampleDistMatrix)), "Female", "Male")

# Extract Dose from sample names:
dose <- sub("-.*", "", colnames(sampleDistMatrix))

# Create a data frame for the column annotation
column_annotation <- data.frame(Sex = sex, Dose = dose)

# Set row names of annotation to match the column names of sampleDistMatrix
rownames(column_annotation) <- colnames(sampleDistMatrix)

pheatmap(sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists, annotation_col = column_annotation,
  col = colors
)

# clustering Genes

topVarGenes <- head(order(rowVars(assay(vsd_corrected)), decreasing = TRUE), 30)

mat <- assay(vsd_corrected)[topVarGenes, ]

mat <- mat - rowMeans(mat)

anno <- as.data.frame(colData(vsd_corrected)[, c("Sex", "Dose", "Donors")])

genes <- rownames(mat)

genes <- mapIds(org.Hs.eg.db,
  keys = genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

rownames(mat) <- genes

pheatmap(mat, annotation_col = anno, cluster_cols = F)

```

# Annotation

```{r}

anno_df <- get_annotation_orgdb(
  dds = dds,
  orgdb_species = "org.Hs.eg.db",
  idtype = "ENSEMBL"
)

```

# Differential Expression Analysis

```{r}

dds$Donors<- factor(dds$Donors)

dds$Dose<- factor(dds$Dose)

design(dds) <- ~ Donors + Dose

dds <- DESeq(dds)


resultsNames(dds)

results_LR<- results(dds, name = "Dose_4Gy_vs_0Gy", alpha = 0.05)

summary(results_LR)

results_LR_shrink<- lfcShrink(dds, res = results_LR, coef = 'Dose_4Gy_vs_0Gy', type="apeglm")

results_LR_shrink_DEs <- deresult_to_df(results_LR_shrink, FDR = 0.05)


View(results_LR_shrink_DEs)

genes <- rownames(results_LR_shrink_DEs)

genes <- mapIds(
  org.Hs.eg.db,
  keys = genes,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)


results_LR_shrink_DEs$SYMBOL <- genes

```

# Volcano plot

```{r}

volcPlot <- de_volcano(
  res_de = results_LR_shrink,
  mapping = "org.Hs.eg.db",
  labeled_genes = 50,
  logfc_cutoff = 1
)
volcPlot

```

## Creating gene plotting function

```{r}

plot_gene <- function (dds, gene, intgroup = "condition", assay = "counts", 
                       annotation_obj = NULL, normalized = TRUE, transform = TRUE, 
                       labels_display = FALSE, labels_repel = TRUE, plot_type = "auto", 
                       return_data = FALSE, color_by = "Donors", gtl = NULL) 
{
  plot_type <- match.arg(plot_type, c("auto", "jitteronly", 
                                      "boxplot", "violin", "sina"))
  
  if (!is.null(gtl)) {
    checkup_gtl(gtl)
    dds <- gtl$dds
    res_de <- gtl$res_de
    res_enrich <- gtl$res_enrich
    annotation_obj <- gtl$annotation_obj
  }

  if (!(all(intgroup %in% colnames(colData(dds))))) {
    stop("`intgroup` not found in the colData slot of the dds object", 
         "\nPlease specify one of the following: \n", paste0(colnames(colData(dds)), 
                                                             collapse = ", "))
  }
  
  df <- get_expression_values(dds = dds, gene = gene, intgroup = intgroup, 
                              assay = assay, normalized = normalized)
  df$sample_id <- rownames(df)
  
  # Explicitly add the Donors column from colData(dds) if it exists
  if ("Donors" %in% colnames(colData(dds))) {
    df$Donors <- colData(dds)[, "Donors"]
  } else {
    stop("The `Donors` column is missing in the colData of the dds object.")
  }
  
  if (!is.null(annotation_obj)) {
    genesymbol <- annotation_obj$gene_name[match(gene, annotation_obj$gene_id)]
  } else {
    genesymbol <- ""
  }
  
  onlyfactors <- df[, match(intgroup, colnames(df))]
  df$plotby <- interaction(onlyfactors)
  min_by_groups <- min(table(df$plotby))
  
  if (return_data) {
    return(df)
  }

  p <- ggplot(df, aes(x = .data$plotby, y = .data$exp_value, col = .data[[color_by]])) + 
    scale_x_discrete(name = "") + 
    scale_color_discrete(name = "Experimental\ngroup") + 
    theme_bw()

  # Add error bars for standard deviation
  #p <- p + stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.2, color = "black")

  jit_pos <- position_jitter(width = 0.2, height = 0, seed = 42)
  
  if (plot_type == "jitteronly" || (plot_type == "auto" & min_by_groups <= 3)) {
    p <- p + geom_point(position = jit_pos)
  } else if (plot_type == "boxplot" || (plot_type == "auto" & (min_by_groups > 3 & min_by_groups < 10))) {
    p <- p + geom_boxplot(outlier.shape = NA) + geom_point(position = jit_pos)
  } else if (plot_type == "violin" || (plot_type == "auto" & (min_by_groups >= 11 & min_by_groups < 40))) {
    p <- p + geom_violin() + geom_point(position = jit_pos) + 
      stat_summary(fun = median, fun.min = median, fun.max = median, geom = "crossbar", width = 0.3)
  } else if (plot_type == "sina" || (plot_type == "auto" & (min_by_groups >= 40))) {
    p <- p + ggforce::geom_sina() + stat_summary(fun = median, 
                                                 fun.min = median, fun.max = median, geom = "crossbar", width = 0.3)
  }

  # Remove sample labels from plot by setting `labels_display = FALSE`
  if (labels_display) {
    if (labels_repel) {
      p <- p + ggrepel::geom_text_repel(aes(label = .data$sample_id), 
                                        min.segment.length = 0, position = jit_pos)
    } else {
      p <- p + geom_text(aes(label = .data$sample_id), hjust = -0.1, vjust = 0.1, position = jit_pos)
    }
  }
  
  # Add grey line representing the mean across all groups
  p <- p + stat_summary(fun = mean, geom = "line", aes(group = 1), 
                        color = "grey80", linewidth = 0.5)

  y_label <- if (assay == "counts" & normalized) {
    "Normalized counts"
  } else if (assay == "counts" & !normalized) {
    "Counts"
  } else if (assay == "abundance") {
    "TPM - Transcripts Per Million"
  } else {
    assay
  }
  
  if (transform) {
    p <- p + scale_y_log10(name = paste0(y_label, " (log10 scale)"))
  } else {
    p <- p + scale_y_continuous(name = y_label)
  }

  if (!is.null(annotation_obj)) {
    p <- p + labs(title = paste0(genesymbol, " - ", gene))
  } else {
    p <- p + labs(title = paste0(gene))
  }

  return(p)
}

```

## Checking the gene pattern across doses


```{r}

# EX--> checking FDXR gene

plot_gene(
  dds= dds,
  gene = "ENSG00000161513",
  intgroup = "Dose",
  annotation_obj = anno_df)

```

# Functional Enrichment Analysis

```{r}
expressedInAssay <- (rowSums(assay(dds)) > 0)
geneUniverseExprENS <- rownames(dds)[expressedInAssay]
geneUniverseExpr <- anno_df$gene_name[match(geneUniverseExprENS, anno_df$gene_id)]

```

```{r}

res_enrich<- run_cluPro(
      de_genes = results_LR_shrink_DEs$SYMBOL,
      res_de = results_LR,
      de_container = dds,
      keyType = "SYMBOL",
      mapping = "org.Hs.eg.db",
      ont = "BP",
      pAdjustMethod = "BH",
      bg_genes = geneUniverseExpr,
      FDR_threshold = 0.05,
    )

View(res_enrich@result)

```

## Visualizing top enriched pathways

```{r}

barplot(res_enrich, showCategory=20)

```

## Creating GeneTonic object

```{r}

gtl_pilot_LR <- GeneTonicList(
  dds = dds,
  res_de = results_LR,
  res_enrich = shake_enrichResult(res_enrich),
  annotation_obj = anno_df
)

```

## Running GeneTonic

```{r, eval=FALSE}
GeneTonic(gtl = gtl_pilot_LR)
```

## GeneTonic Heatmap

```{r}
scored_gtl_mat <- GeneTonic::gs_scores(se = vsd_corrected, gtl = gtl_pilot_LR)

anno_full <- as.data.frame(colData(vsd_phybion_LR)[, c("Sex", "Dose")])

column_annotation <- HeatmapAnnotation(
  Sex = anno_full$Sex,
  Dose = anno_full$Dose)

Heatmap(scored_gtl_mat[1:30,],
  name = "Expression",
  top_annotation = column_annotation,
  cluster_columns = F, # Optional: Cluster columns
  cluster_rows = F, # Optional: Cluster rows
  show_row_names = TRUE,
  show_column_names = TRUE, row_dend_side = "left",
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 7),
  width = unit(8, "cm"),
  heatmap_height = unit(10, "cm"),
  column_names_gp = gpar(fontsize = 6),
  column_names_rot = 360, 
  column_title = "Donors", 
  column_title_side = "bottom"
)

# OR

GeneTonic::gs_scoresheat(scored_gtl_mat, n_gs = 30, cluster_cols = FALSE)

```

# Session information

```{r sessioninfo}
sessionInfo()
```
